---
title: "Analysis of GSE26145"
author: "T. Schätzl and M. Kohl"
date: "May 2023"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
## setup for the chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)
Sys.setlocale("LC_ALL", "English")
```


# Import of data

The data are included in dataset GSE26145 of Gene Expression Omnibus:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26145

The tar-file was unzipped into a subfolder called GSE26145_RAW. 
We removed the disease control.

We import the data from the CEL-files using Bioconductor package oligo.

```{r}
library(oligo)
celFiles <- list.files("./GSE26145_RAW", full.names = TRUE)
rawData <- read.celfiles(celFiles)
rawData
```


# Quality control

## QC report of package arrayQualityMetrics

```{r, eval=FALSE}
library(arrayQualityMetrics)
arrayQualityMetrics(rawData, outdir = "QC", force = TRUE, do.logtransform = TRUE)
```

## Similarity map

```{r}
library(MKomics)
COR <- cor(exprs(rawData), method = "spearman")
corPlot2(COR, minCor = min(COR), fontsize.axis = 8, 
         title = "Spearman correlation")
```


# Preprocessing

We use RMA to preprocess the data.

```{r, cache = TRUE}
normData <- rma(rawData)
normData
```


# Quality control of preprocessed data

## Mean-SD plot

```{r}
library(vsn)
meanSdPlot(normData)
```

## Boxplot

```{r}
## range = 0: whiskers till most extreme data points
boxplot(exprs(normData), range = 0, main = "RMA", las = 2, 
        cex.axis = 0.75)
```

## Similarity map

```{r}
COR.rma <- cor(exprs(normData), method = "spearman")
corPlot2(COR.rma, minCor = min(COR.rma), fontsize.axis = 8, 
         title = "Spearman correlation")
```

## Import of sample information

```{r}
SampleInfo <- read.csv(file = "GSE26145_SampleInfo.csv", na.strings = "N/A")
SampleInfo
```

Due to different scan dates between FSHD patients and controls (see "GSE26145_SampleInfo.csv"), there might be a batch effect. Batch effects cannot be removed, as we do not know due to the experimental design whether the occurring differences are due to different scan dates or actual differences between FSHD patients and controls.

## Saving of preprocessed data

```{r}
ExportData <- t(exprs(normData))
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE26145data.rds")
```


# Statistical analysis

## Moderated t-test by packages limma and MKomics

```{r}
library(limma)
library(MKomics)
res <- mod.t.test (x = exprs(normData), group = SampleInfo$group)
colSums(res[,c(5,6)] < 0.05)
# no results for adj. p value < 5%
```


# Used software

```{r}
sessionInfo()
```
