---
title: "Analysis of EGAD00001008337"
author: "T. Schaetzl and M. Kohl"
date: "August 2022"
output:
  rmarkdown::html_document:
    theme: united
    highlight: tango
    toc: true
---

# Loading of required packages

```{r}
library(tximport)
library(MKomics)
library(vsn)
library(limma)
library(MKinfer)
library(VennDiagram)
library(edgeR)
library(GO.db)
library(org.Hs.eg.db)
```


# Import of data

```{r}
SampleInfo <- read.csv("EGAD00001008337_SampleInfo.csv", na.strings = "N/A")
SampleInfo
cts <- read.delim(file='01_Data --- EGAS00001005782 --- RawCounts_Exprgenesonly.txt',row.names = 1)
# Filtering has already been done
cts <- as.matrix(cts)
# Removing mosaic samples
cts <- cts[,SampleInfo$Repeat_units != "Mosaic_2units"]
SampleInfo <- SampleInfo[SampleInfo$Repeat_units != "Mosaic_2units",]
#Removing asymptomatic samples
cts <- cts[,SampleInfo$SampleID != "FSHD_15_VL"]
SampleInfo <- SampleInfo[SampleInfo$SampleID != "FSHD_15_VL",]
cts <- cts[,SampleInfo$SampleID != "FSHD_23_VL"]
SampleInfo <- SampleInfo[SampleInfo$SampleID != "FSHD_23_VL",]
sort(colSums(cts)/1e6)
```


# Quality control of raw data

```{r}
cts0 <- cts
cts0[cts0 == 0] <- min(cts0[cts0 != 0])
boxplot(cts0, log = "y", range = 0, las = 2)
```

```{r}
Cor <- cor(cts, method = "spearman")
corPlot2(Cor, minCor = min(Cor), 
         #labels = paste(SampleInfo$subject),
         title = "Spearman correlation")
```


# Preprocessing

```{r}
dge <- DGEList(counts=cts)
group <- SampleInfo$group
logCPM <- cpm(dge, log=TRUE, prior.count=3)
```


## Quality control of filtered data

```{r}
boxplot(logCPM, range = 0, las = 2)
```

```{r}
Cor <- cor(logCPM, method = "spearman")
corPlot2(Cor, minCor = min(Cor), 
         #labels=paste(SampleInfo$subject),
         title = "Spearman correlation")
```

```{r}
meanSdPlot(logCPM)
```

```{r}
library(org.Hs.eg.db)
ensemblIDs <- rownames(logCPM)
entrezIDs <- mapIds(org.Hs.eg.db, keys = ensemblIDs, column = "ENTREZID", 
                    keytype = "ENSEMBL", multiVals = "first")
symbols <- mapIds(org.Hs.eg.db, keys = ensemblIDs
                  , column = "SYMBOL", 
                    keytype = "ENSEMBL", multiVals = "first")
```

```{r}
ExportData <- t(logCPM)
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "EGAD00001008337data.rds")
```


# Statistical Analysis

## Gene list

```{r}
library(openxlsx)
GeneSel <- read.xlsx("FSHDlist.xlsx")
GeneSel$ENSEMBL <- sapply(strsplit(GeneSel$gene_id, "\\."), "[", 1)
dim(GeneSel)
```

```{r}
logCPM.Genesel <- logCPM[ensemblIDs %in% GeneSel$ENSEMBL,]
GeneSel[!(GeneSel$ENSEMBL %in% ensemblIDs),]
```

```{r}
library(org.Hs.eg.db)
ensemblIDs.Genesel <- rownames(logCPM.Genesel)
entrezIDs.Genesel <- mapIds(org.Hs.eg.db, keys = ensemblIDs.Genesel, column = "ENTREZID", 
                    keytype = "ENSEMBL", multiVals = "first")
symbols.Genesel <- mapIds(org.Hs.eg.db, keys = ensemblIDs.Genesel, column = "SYMBOL", 
                    keytype = "ENSEMBL", multiVals = "first")
```

```{r}
library(MKomics)
design <- model.matrix(~group)
fit <- lmFit(logCPM.Genesel, design)
fit <- eBayes(fit)
res.GeneSel <- topTable(fit, number = Inf)
sum(res.GeneSel$P.Value < 0.05)
# 63 results
sort(unique(symbols.Genesel[res.GeneSel$P.Value < 0.05]))
sum(res.GeneSel$adj.P.Val < 0.05)
# 41 results
sort(unique(symbols.Genesel[res.GeneSel$adj.P.Val < 0.05]))
```


### Export of Results, Gene list

```{r}
res.GeneSel.EGAD00001008337 <- as.data.frame(res.GeneSel)
res.GeneSel.EGAD00001008337$ENSEMBLE <- ensemblIDs.Genesel
res.GeneSel.EGAD00001008337 <- res.GeneSel.EGAD00001008337[order(res.GeneSel.EGAD00001008337$P.Val),]
write.csv(res.GeneSel.EGAD00001008337, file = "ResultsGeneSel_EGAD00001008337.csv", row.names = FALSE)
save(res.GeneSel.EGAD00001008337, file = "ResultsGeneSel_EGAD00001008337.RData")
```


## All genes

```{r}
library(MKomics)
fit <- lmFit(logCPM, design)
fit <- eBayes(fit)
res <- topTable(fit, number = Inf)
sum(res$P.Value < 0.05)
sort(unique(symbols[res$P.Value < 0.05]))
sum(res$adj.P.Val < 0.05)
# 7326 results for adj. p value < 5%
sort(unique(symbols[res$adj.P.Val < 0.05]))
```


### Export of Results

```{r}
res.EGAD00001008337 <- as.data.frame(res)
res.EGAD00001008337$ENSEMBLE <- ensemblIDs
res.EGAD00001008337 <- res.EGAD00001008337[order(res.EGAD00001008337$adj.P.Val),]
write.csv(res.EGAD00001008337, file = "ResultsEGAD00001008337.csv", row.names = FALSE)
save(res.EGAD00001008337, file = "ResultsEGAD00001008337.RData")
```


## Heatmap 

```{r}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
## Data selection
sel <-res$adj.P.Val < 0.05
selData <- logCPM[sel,]
selData <- selData - rowMeans(selData, na.rm = TRUE)
colnames(selData) <- paste(SampleInfo$SampleID)
## Color definition
col1 <- rev(brewer.pal(n = 8, name = "RdYlBu"))
col2 <- brewer.pal(n = 3, name = "Set1")[c(3,1)]
col3 <- brewer.pal(n = 3, name = "Set2")[c(3,1)]
col4 <- brewer.pal(n = 3, name = "Dark2")[c(2,4)]
mycol <- colorRamp2(seq(-1, 1, length = 128), 
                    colorRampPalette(col1)(128))
## Annotation
df <- data.frame(group = SampleInfo$group)
ha <- HeatmapAnnotation(df = df,
                        col = list(group = c("CTRL" = col2[1],
                                              "FSHD" = col2[2])))
## Drawing the heatmap
pdf(file="newPlot.pdf", height = 15, width = 8)
print(
Heatmap(selData, col = mycol, name = "log2-FC", show_row_names = FALSE,
        use_raster = FALSE,
        top_annotation = ha, show_column_names = TRUE,
        column_names_gp = gpar(fontsize = 8),
        clustering_distance_columns = "euclidean",
        show_row_dend = TRUE, km = 1,
        cluster_columns = TRUE, show_column_dend = TRUE,
        column_title = "p value < 0.05 and |log2-FC| > log2(1.5)",
        show_heatmap_legend = TRUE))
dev.off()
```


### Enrichment Analysis

```{r}

GeneID.UP <- entrezIDs[(res$logFC > 0 & 
                        res$adj.P.Val < 0.05)]
GeneID.DN <- entrezIDs[(res$logFC > 0 & 
                        res$adj.P.Val < 0.05)]
```


#### Calculation of enrichment for GO

```{r}
library(GO.db)
go <- goana(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:20,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:20,]
```


### Export of Results: GO

```{r}
go.up.EGAD00001008337 <- go.up
go.up.EGAD00001008337 <- go.up.EGAD00001008337[order(go.up.EGAD00001008337$P.Up),]
write.csv(go.up.EGAD00001008337, file = "GO.UP_EGAD00001008337.csv", row.names = FALSE)
save(go.up.EGAD00001008337, file = "GO.UP_EGAD00001008337.RData")
```

```{r}
go.dn.EGAD00001008337 <- go.dn
go.dn.EGAD00001008337 <- go.dn.EGAD00001008337[order(go.dn.EGAD00001008337$P.Down),]
write.csv(go.dn.EGAD00001008337, file = "GO.DN_EGAD00001008337.csv", row.names = FALSE)
save(go.dn.EGAD00001008337, file = "GO.DN_EGAD00001008337.RData")
```


#### Calculation of enrichment for KEGG

```{r}
kg <- kegga(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
kegg.up <- topKEGG(kg, sort = "up", number = Inf)
kegg.up[1:20,]
kegg.dn <- topKEGG(kg, sort = "down", number = Inf)
kegg.dn[1:20,]
```


### Export of Results: KEGG

```{r}
kegg.up.EGAD00001008337 <- kegg.up
kegg.up.EGAD00001008337 <- kegg.up.EGAD00001008337[order(kegg.up.EGAD00001008337$P.Up),]
write.csv(kegg.up.EGAD00001008337, file = "KEGG.UP_EGAD00001008337.csv", row.names = FALSE)
save(kegg.up.EGAD00001008337, file = "KEGG.UP_EGAD00001008337.RData")
```

```{r}
kegg.dn.EGAD00001008337 <- kegg.dn
kegg.dn.EGAD00001008337 <- kegg.dn.EGAD00001008337[order(kegg.dn.EGAD00001008337$P.Down),]
write.csv(kegg.dn.EGAD00001008337, file = "KEGG.DN_EGAD00001008337.csv", row.names = FALSE)
save(kegg.dn.EGAD00001008337, file = "KEGG.DN_EGAD00001008337.RData")
```


# Used software

```{r}
sessionInfo()
```
