---
title: "Analysis of GSE123468"
author: "T. Schaetzl and M. Kohl"
date: "August 2022"
output:
  rmarkdown::html_document:
    theme: united
    highlight: tango
    toc: true
---

# Loading of required packages

```{r}
library(tximport)
library(MKomics)
library(vsn)
library(limma)
library(MKinfer)
library(VennDiagram)
library(edgeR)
library(GO.db)
library(org.Hs.eg.db)
```


# Import of Salmon Quantification data

```{r}
SampleInfo <- read.csv("GSE123468_SampleInfo.csv")
SampleInfo
txi <- readRDS(file = "SRP172788.rds")
head(txi$counts)
cts <- txi$counts
cts <- cts[,order(colnames(cts))]
sort(colSums(cts)/1e6)
```


# Quality control of raw data

```{r}
cts0 <- cts
cts0[cts0 == 0] <- min(cts0[cts0 != 0])
boxplot(cts0, log = "y", range = 0, las = 2)
```

```{r}
Cor <- cor(cts, method = "spearman")
corPlot2(Cor, minCor = min(Cor), 
         labels = paste(SampleInfo$subject, SampleInfo$celltype),
         title = "Spearman correlation")
```


# Preprocessing

```{r}
dge <- DGEList(counts=cts)
group <- factor(SampleInfo$group)
celltype <- factor(SampleInfo$celltype)
family <- factor(SampleInfo$family)
design <- model.matrix(~ group * celltype * family)
keep <- filterByExpr(dge, design)
dge <- dge[keep,,keep.lib.sizes=FALSE]
dge <- calcNormFactors(dge)
logCPM <- cpm(dge, log=TRUE, prior.count=3)
```


## Quality control of filtered data

```{r}
boxplot(logCPM, range = 0, las = 2)
```

```{r}
Cor <- cor(logCPM, method = "spearman")
corPlot2(Cor, minCor = min(Cor), 
         labels = paste(SampleInfo$subject, SampleInfo$celltype),
         title = "Spearman correlation")
```

```{r}
meanSdPlot(logCPM)
```

```{r}
library(org.Hs.eg.db)
symbols <- sapply(strsplit(rownames(logCPM), "\\|"), "[", 6)
ensemblIDs <- sapply(strsplit(rownames(logCPM), "\\|"), "[", 2)
ensemblIDs <- sapply(strsplit(ensemblIDs, "\\."), "[", 1)
entrezIDs <- mapIds(org.Hs.eg.db, keys = ensemblIDs, column = "ENTREZID", 
                    keytype = "ENSEMBL", multiVals = "first")
```

```{r}
ExportData <- t(logCPM)
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE123468data.rds")
```


# Statistical Analysis

## Gene list

```{r}
library(openxlsx)
GeneSel <- read.xlsx("FSHDlist.xlsx")
GeneSel$ENSEMBL <- sapply(strsplit(GeneSel$gene_id, "\\."), "[", 1)
dim(GeneSel)
```

```{r}
logCPM.Genesel <- logCPM[ensemblIDs %in% GeneSel$ENSEMBL,]
GeneSel[!(GeneSel$ENSEMBL %in% ensemblIDs),]
```

```{r}
library(org.Hs.eg.db)
symbols.Genesel <- sapply(strsplit(rownames(logCPM.Genesel), "\\|"), "[", 6)
ensemblIDs.Genesel <- sapply(strsplit(rownames(logCPM.Genesel), "\\|"), "[", 2)
ensemblIDs.Genesel <- sapply(strsplit(ensemblIDs.Genesel, "\\."), "[", 1)
entrezIDs.Genesel <- mapIds(org.Hs.eg.db, keys = ensemblIDs.Genesel, column = "ENTREZID", 
                    keytype = "ENSEMBL", multiVals = "first")
```

```{r}
###########################################################
## family 12
###########################################################
group12 <- group[family == "12"]
celltype12 <- celltype[family == "12"]
design12 <- model.matrix(~ group12 * celltype12)
fit12 <- lmFit(logCPM.Genesel[,family == "12"], design12)
fit12 <- eBayes(fit12)

## group
resgroup12 <- topTable(fit12, coef="group12FSHD", number = Inf)
sum(resgroup12$P.Value < 0.05)
# 43 results
sort(unique(symbols.Genesel[resgroup12$P.Val < 0.05]))
sum(resgroup12$adj.P.Val < 0.05)
# 24 results
sort(unique(symbols.Genesel[resgroup12$adj.P.Val < 0.05]))


## celltype
rescelltype12 <- topTable(fit12, coef="celltype12MT(T8)", number = Inf)
sum(rescelltype12$P.Value < 0.05)
# 6 results
sort(unique(symbols.Genesel[rescelltype12$P.Val < 0.05]))
sum(rescelltype12$adj.P.Val < 0.05)
# 1 result
sort(unique(symbols.Genesel[rescelltype12$adj.P.Val < 0.05]))

## group:celltype
resgroup.celltype12 <- topTable(fit12, coef="group12FSHD:celltype12MT(T8)", 
                                number = Inf)
sum(resgroup.celltype12$P.Value < 0.05)
# 10 results
sort(unique(symbols.Genesel[resgroup.celltype12$P.Val < 0.05]))
sum(resgroup.celltype12$adj.P.Val < 0.05)
# no results

###########################################################
## family 16
###########################################################
group16 <- group[family == "16"]
celltype16 <- celltype[family == "16"]
design16 <- model.matrix(~ group16 * celltype16)
fit16 <- lmFit(logCPM.Genesel[,family == "16"], design16)
fit16 <- eBayes(fit16)

## group
resgroup16 <- topTable(fit16, coef="group16FSHD", number = Inf)
sum(resgroup16$P.Value < 0.05)
# 67 results
sort(unique(symbols.Genesel[resgroup16$P.Val < 0.05]))
sum(resgroup16$adj.P.Val < 0.05)
# 52 results
sort(unique(symbols.Genesel[resgroup16$adj.P.Val < 0.05]))

## celltype
rescelltype16 <- topTable(fit16, coef="celltype16MT(T8)", number = Inf)
sum(rescelltype16$P.Value < 0.05)
# 8 results
sort(unique(symbols.Genesel[rescelltype16$P.Val < 0.05]))
sum(rescelltype16$adj.P.Val < 0.05)
# no results

## group:celltype
resgroup.celltype16 <- topTable(fit16, coef="group16FSHD:celltype16MT(T8)", 
                                number = Inf)
sum(resgroup.celltype16$P.Value < 0.05)
# 11 results
sort(unique(symbols.Genesel[resgroup.celltype16$P.Val < 0.05]))
sum(resgroup.celltype16$adj.P.Val < 0.05)
# no results

###########################################################
## families 12 & 16
###########################################################
##group
sort(unique(union(symbols.Genesel[resgroup12$P.Val < 0.05], 
          symbols.Genesel[resgroup16$P.Val < 0.05])))
sort(unique(union(symbols.Genesel[resgroup12$adj.P.Val < 0.05], 
          symbols.Genesel[resgroup16$adj.P.Val < 0.05])))

## celltype
sort(unique(union(symbols.Genesel[rescelltype12$P.Val < 0.05],
          symbols.Genesel[rescelltype16$P.Val < 0.05])))
sort(unique(union(symbols.Genesel[rescelltype12$adj.P.Val < 0.05],
          symbols.Genesel[rescelltype16$adj.P.Val < 0.05])))

## group:celltype
sort(unique(union(symbols.Genesel[resgroup.celltype12$P.Val < 0.05],
          symbols.Genesel[resgroup.celltype16$P.Val < 0.05])))
sort(unique(union(symbols.Genesel[resgroup.celltype12$adj.P.Val < 0.05],
          symbols.Genesel[resgroup.celltype16$adj.P.Val < 0.05])))
```


## Export of Results Fam12

```{r}
temp1 <- resgroup12
colnames(temp1) <- paste0("group12: ", colnames(resgroup12))
temp2 <- rescelltype12
colnames(temp2) <- paste0("celltype12: ", colnames(rescelltype12))
temp3 <- resgroup.celltype12
colnames(temp3) <- paste0("group:celltype12: ", 
                                      colnames(resgroup.celltype12))
temp1.ID <- sapply(strsplit(rownames(temp1), "\\|"), "[", 2)
temp1.ID <- sapply(strsplit(temp1.ID, "\\."), "[", 1)
temp2.ID <- sapply(strsplit(rownames(temp2), "\\|"), "[", 2)
temp2.ID <- sapply(strsplit(temp2.ID, "\\."), "[", 1)
temp3.ID <- sapply(strsplit(rownames(temp3), "\\|"), "[", 2)
temp3.ID <- sapply(strsplit(temp3.ID, "\\."), "[", 1)
res.GeneSel.GSE123468_12 <- cbind(temp1[order(temp1.ID),], temp2[order(temp2.ID),], temp3[order(temp3.ID),])
res.GeneSel.GSE123468_12$ENSEMBLE <- sort(temp1.ID)
rm(temp1,temp2, temp3)
write.csv(res.GeneSel.GSE123468_12, file = "ResultsGenesel_GSE123468_Fam12.csv")
save(res.GeneSel.GSE123468_12, file = "ResultsGenesel_GSE123468_Fam12.RData")
```


## Export of Results Fam16

```{r}
temp1 <- resgroup16
colnames(temp1) <- paste0("group16: ", colnames(resgroup16))
temp2 <- rescelltype16
colnames(temp2) <- paste0("celltype16: ", colnames(rescelltype16))
temp3 <- resgroup.celltype16
colnames(temp3) <- paste0("group:celltype16: ", 
                                      colnames(resgroup.celltype16))
temp1.ID <- sapply(strsplit(rownames(temp1), "\\|"), "[", 2)
temp1.ID <- sapply(strsplit(temp1.ID, "\\."), "[", 1)
temp2.ID <- sapply(strsplit(rownames(temp2), "\\|"), "[", 2)
temp2.ID <- sapply(strsplit(temp2.ID, "\\."), "[", 1)
temp3.ID <- sapply(strsplit(rownames(temp3), "\\|"), "[", 2)
temp3.ID <- sapply(strsplit(temp3.ID, "\\."), "[", 1)
res.GeneSel.GSE123468_16 <- cbind(temp1[order(temp1.ID),], temp2[order(temp2.ID),], temp3[order(temp3.ID),])
res.GeneSel.GSE123468_16$ENSEMBLE <- sort(temp1.ID)
rm(temp1,temp2, temp3)
write.csv(res.GeneSel.GSE123468_16, file = "ResultsGenesel_GSE123468_Fam16.csv")
save(res.GeneSel.GSE123468_16, file = "ResultsGenesel_GSE123468_Fam16.RData")
```


## All genes

```{r}
###########################################################
## family 12
###########################################################
group12 <- group[family == "12"]
celltype12 <- celltype[family == "12"]
design12 <- model.matrix(~ group12 * celltype12)
fit12 <- lmFit(logCPM[,family == "12"], design12)
fit12 <- eBayes(fit12)

## group
resgroup12 <- topTable(fit12, coef="group12FSHD", number = Inf)
sum(resgroup12$P.Value < 0.05)
sum(resgroup12$adj.P.Val < 0.05)
# 347 results for adj. p value < 5%
sort(unique(symbols[resgroup12$adj.P.Val < 0.05]))

## celltype
rescelltype12 <- topTable(fit12, coef="celltype12MT(T8)", number = Inf)
sum(rescelltype12$P.Value < 0.05)
sum(rescelltype12$adj.P.Val < 0.05)
# 21 results for adj. p value < 5%
sort(unique(symbols[rescelltype12$adj.P.Val < 0.05]))

## group:celltype
resgroup.celltype12 <- topTable(fit12, coef="group12FSHD:celltype12MT(T8)", 
                                number = Inf)
sum(resgroup.celltype12$P.Value < 0.05)
sum(resgroup.celltype12$adj.P.Val < 0.05)
# 28 results for adj. p value < 5%
sort(unique(symbols[resgroup.celltype12$adj.P.Val < 0.05]))

###########################################################
## family 16
###########################################################
group16 <- group[family == "16"]
celltype16 <- celltype[family == "16"]
design16 <- model.matrix(~ group16 * celltype16)
fit16 <- lmFit(logCPM[,family == "16"], design16)
fit16 <- eBayes(fit16)

## group
resgroup16 <- topTable(fit16, coef="group16FSHD", number = Inf)
sum(resgroup16$P.Value < 0.05)
sum(resgroup16$adj.P.Val < 0.05)
# 1579 results for adj. p value < 5%
sort(unique(symbols[resgroup16$adj.P.Val < 0.05]))

## celltype
rescelltype16 <- topTable(fit16, coef="celltype16MT(T8)", number = Inf)
sum(rescelltype16$P.Value < 0.05)
sum(rescelltype16$adj.P.Val < 0.05)
# 36 results for adj. p value < 5%
sort(unique(symbols[rescelltype16$adj.P.Val < 0.05]))

## group:celltype
resgroup.celltype16 <- topTable(fit16, coef="group16FSHD:celltype16MT(T8)", 
                                number = Inf)
sum(resgroup.celltype16$P.Value < 0.05)
sum(resgroup.celltype16$adj.P.Val < 0.05)
# 68 results for adj. p value < 5%
sort(unique(symbols[resgroup.celltype16$adj.P.Val < 0.05]))

###########################################################
## families 12 & 16
###########################################################
##group
sum(resgroup12$adj.P.Val < 0.05, resgroup16$adj.P.Val < 0.05)
sort(unique(union(symbols[resgroup12$adj.P.Val < 0.05],
          symbols[resgroup16$adj.P.Val < 0.05])))
## celltype
sum(rescelltype12$adj.P.Val < 0.05, rescelltype16$adj.P.Val < 0.05)
sort(unique(union(symbols[rescelltype12$adj.P.Val < 0.05],
          symbols[rescelltype16$adj.P.Val < 0.05])))

## group:celltype
sum(resgroup.celltype12$adj.P.Val < 0.05, resgroup.celltype16$adj.P.Val < 0.05)
sort(unique(union(symbols[resgroup.celltype12$adj.P.Val < 0.05],
          symbols[resgroup.celltype16$adj.P.Val < 0.05])))
```


## Export of Results Fam12

```{r}
temp1 <- resgroup12
colnames(temp1) <- paste0("group12: ", colnames(resgroup12))
temp2 <- rescelltype12
colnames(temp2) <- paste0("celltype12: ", colnames(rescelltype12))
temp3 <- resgroup.celltype12
colnames(temp3) <- paste0("group:celltype12: ", 
                                      colnames(resgroup.celltype12))

res.GSE123468_12 <- cbind(temp1, temp2, temp3, ensemblIDs)
rm(temp1,temp2, temp3)
names(res.GSE123468_12)[19] <- "ENSEMBLE"
write.csv(res.GSE123468_12, file = "ResultsGSE123468_Fam12.csv")
save(res.GSE123468_12, file = "ResultsGSE123468_Fam12.RData")
```

## Export of Results Fam16

```{r}
temp1 <- resgroup16
colnames(temp1) <- paste0("group16: ", colnames(resgroup16))
temp2 <- rescelltype16
colnames(temp2) <- paste0("celltype16: ", colnames(rescelltype16))
temp3 <- resgroup.celltype16
colnames(temp3) <- paste0("group:celltype16: ", 
                                      colnames(resgroup.celltype16))
res.GSE123468_16 <- cbind(temp1, temp2, temp3, ensemblIDs)
rm(temp1,temp2, temp3)
names(res.GSE123468_16)[19] <- "ENSEMBLE"
write.csv(res.GSE123468_16, file = "ResultsGSE123468_Fam16.csv")
save(res.GSE123468_16, file = "ResultsGSE123468_Fam16.RData")
```


## Heatmap 

```{r}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
## Data selection
sel <- p.adjust(fit12$F.p.value, "fdr") < 0.05
selData <- logCPM[sel,]
selData <- selData - rowMeans(selData, na.rm = TRUE)
colnames(selData) <- paste(SampleInfo$subject, SampleInfo$celltype)
## Color definition
col1 <- rev(brewer.pal(n = 8, name = "RdYlBu"))
col2 <- brewer.pal(n = 3, name = "Set1")[c(3,1)]
col3 <- brewer.pal(n = 3, name = "Set2")[c(3,1)]
col4 <- brewer.pal(n = 3, name = "Dark2")[c(2,4)]
mycol <- colorRamp2(seq(-1, 1, length = 128), 
                    colorRampPalette(col1)(128))
## Annotation
df <- data.frame(group = group,
                 celltype = celltype,
                 family = family)
ha <- HeatmapAnnotation(df = df,
                        col = list(group = c("control" = col2[1],
                                              "FSHD" = col2[2]),
                                   celltype = c("MB(T1)" = col3[1],
                                                "MT(T8)" = col3[2]),
                                   family = c("12" = col4[1],
                                              "16" = col4[2])))
## Drawing the heatmap
#Heatmap(selData, col = mycol, name = "log2-FC", show_row_names = FALSE,
#        use_raster = TRUE, raster_device = "CairoPNG",
#        top_annotation = ha, show_column_names = TRUE,
#        column_names_gp = gpar(fontsize = 8),
#        clustering_distance_columns = "pearson",
#        show_row_dend = FALSE, km = 3,
#        cluster_columns = TRUE, show_column_dend = TRUE,
#        column_title = "p value < 0.1 and |log2-FC| > log2(1.5)",
#        show_heatmap_legend = TRUE)
```


## 1.Enrichment Analysis: group


### 1. group, Family 12

```{r}
GeneID.UP <- entrezIDs[(resgroup12$logFC > 0 & 
                        resgroup12$adj.P.Val < 0.05)]
GeneID.DN <- entrezIDs[(resgroup12$logFC < 0 & 
                        resgroup12$adj.P.Val < 0.05)]
```


#### 1. group, Family 12: Calculation of enrichment for GO

```{r}
library(GO.db)
go <- goana(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:20,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:20,]
```


### Export of Results: GO,  1. group, Family 12

```{r}
go.up.GSE123468_12 <- go.up
go.up.GSE123468_12 <- go.up.GSE123468_12[order(go.up.GSE123468_12$P.Up),]
write.csv(go.up.GSE123468_12, file = "GO.UP_GSE123468_groupFam12.csv", row.names = FALSE)
save(go.up.GSE123468_12, file = "GO.UP_GSE123468_groupFam12.RData")
```

```{r}
go.dn.GSE123468_12 <- go.dn
go.dn.GSE123468_12 <- go.dn.GSE123468_12[order(go.dn.GSE123468_12$P.Down),]
write.csv(go.dn.GSE123468_12, file = "GO.DN_GSE123468_groupFam12.csv", row.names = FALSE)
save(go.dn.GSE123468_12, file = "GO.DN_GSE123468_groupFam12.RData")
```


#### 1. group, Family 12: Calculation of enrichment for KEGG

```{r}
kg <- kegga(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
kegg.up <- topKEGG(kg, sort = "up", number = Inf)
kegg.up[1:20,]
kegg.dn <- topKEGG(kg, sort = "down", number = Inf)
kegg.dn[1:20,]
```


### Export of Results: KEGG, 1. group, Family 12

```{r}
kegg.up.GSE123468_12 <- kegg.up
kegg.up.GSE123468_12 <- kegg.up.GSE123468_12[order(kegg.up.GSE123468_12$P.Up),]
write.csv(kegg.up.GSE123468_12, file = "KEGG.UP_GSE123468_groupFam12.csv", row.names = FALSE)
save(kegg.up.GSE123468_12, file = "KEGG.UP_GSE123468_groupFam12.RData")
```

```{r}
kegg.dn.GSE123468_12 <- kegg.dn
kegg.dn.GSE123468_12 <- kegg.dn.GSE123468_12[order(kegg.dn.GSE123468_12$P.Down),]
write.csv(kegg.dn.GSE123468_12, file = "KEGG.DN_GSE123468_groupFam12.csv", row.names = FALSE)
save(kegg.dn.GSE123468_12, file = "KEGG.DN_GSE123468_groupFam12.RData")
```


### 1. group, Family 16

```{r}
GeneID.UP <- entrezIDs[(resgroup16$logFC > 0 & 
                        resgroup16$adj.P.Val < 0.05)]
GeneID.DN <- entrezIDs[(resgroup16$logFC < 0 & 
                        resgroup16$adj.P.Val < 0.05)]
```


#### 1. group, Family 16: Calculation of enrichment for GO

```{r}
library(GO.db)
go <- goana(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:20,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:20,]
```


### Export of Results: GO,  1. group, Family 16

```{r}
go.up.GSE123468_16 <- go.up
go.up.GSE123468_16 <- go.up.GSE123468_16[order(go.up.GSE123468_16$P.Up),]
write.csv(go.up.GSE123468_16, file = "GO.UP_GSE123468_groupFam16.csv", row.names = FALSE)
save(go.up.GSE123468_16, file = "GO.UP_GSE123468_groupFam16.RData")
```

```{r}
go.dn.GSE123468_16 <- go.dn
go.dn.GSE123468_16 <- go.dn.GSE123468_16[order(go.dn.GSE123468_16$P.Down),]
write.csv(go.dn.GSE123468_16, file = "GO.DN_GSE123468_groupFam16.csv", row.names = FALSE)
save(go.dn.GSE123468_16, file = "GO.DN_GSE123468_groupFam16.RData")
```


#### 1. group, Family 16: Calculation of enrichment for KEGG

```{r}
kg <- kegga(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
kegg.up <- topKEGG(kg, sort = "up", number = Inf)
kegg.up[1:20,]
kegg.dn <- topKEGG(kg, sort = "down", number = Inf)
kegg.dn[1:20,]
```


### Export of Results: KEGG, 1. group, Family 16

```{r}
kegg.up.GSE123468_16 <- kegg.up
kegg.up.GSE123468_16 <- kegg.up.GSE123468_16[order(kegg.up.GSE123468_16$P.Up),]
write.csv(kegg.up.GSE123468_16, file = "KEGG.UP_GSE123468_groupFam16.csv", row.names = FALSE)
save(kegg.up.GSE123468_16, file = "KEGG.UP_GSE123468_groupFam16.RData")
```

```{r}
kegg.dn.GSE123468_16 <- kegg.dn
kegg.dn.GSE123468_16 <- kegg.dn.GSE123468_16[order(kegg.dn.GSE123468_16$P.Down),]
write.csv(kegg.dn.GSE123468_16, file = "KEGG.DN_GSE123468_groupFam16.csv", row.names = FALSE)
save(kegg.dn.GSE123468_16, file = "KEGG.DN_GSE123468_groupFam16.RData")
```


### 1. group, Families 12 & 16

```{r}
GeneID.UP.12 <- entrezIDs[(resgroup12$logFC > 0 & 
                        resgroup12$adj.P.Val < 0.05)]
GeneID.UP.16 <- entrezIDs[(resgroup16$logFC > 0 & 
                        resgroup16$adj.P.Val < 0.05)]
GeneID.UP <- union(GeneID.UP.12, GeneID.UP.16)

GeneID.DN.12 <- entrezIDs[(resgroup12$logFC < 0 & 
                        resgroup12$adj.P.Val < 0.05)]
GeneID.DN.16 <- entrezIDs[(resgroup16$logFC < 0 & 
                        resgroup16$adj.P.Val < 0.05)]
GeneID.DN <- union(GeneID.DN.12, GeneID.DN.16)
```


#### 1. group, Families 12 & 16: Calculation of enrichment for GO

```{r}
library(GO.db)
go <- goana(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:50,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:50,]
```


### Export of Results: GO,  1. group, Families 12&16

```{r}
go.up.GSE123468_12and16 <- go.up
go.up.GSE123468_12and16 <- go.up.GSE123468_12and16[order(go.up.GSE123468_12and16$P.Up),]
write.csv(go.up.GSE123468_12and16, file = "GO.UP_GSE123468_groupFam12&16.csv", row.names = FALSE)
save(go.up.GSE123468_12and16, file = "GO.UP_GSE123468_groupFam12&16.RData")
```

```{r}
go.dn.GSE123468_12and16 <- go.dn
go.dn.GSE123468_12and16 <- go.dn.GSE123468_12and16[order(go.dn.GSE123468_12and16$P.Down),]
write.csv(go.dn.GSE123468_12and16, file = "GO.DN_GSE123468_groupFam12&16.csv", row.names = FALSE)
save(go.dn.GSE123468_12and16, file = "GO.DN_GSE123468_groupFam12&16.RData")
```


#### 1. group, Families 12 & 16: Calculation of enrichment for KEGG

```{r}
kg <- kegga(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
kegg.up <- topKEGG(kg, sort = "up", number = Inf)
kegg.up[1:10,]
kegg.dn <- topKEGG(kg, sort = "down", number = Inf)
kegg.dn[1:10,]
```


### Export of Results: KEGG, 1. group, Families 12 & 16

```{r}
kegg.up.GSE123468_12and16 <- kegg.up
kegg.up.GSE123468_12and16 <- kegg.up.GSE123468_12and16[order(kegg.up.GSE123468_12and16$P.Up),]
write.csv(kegg.up.GSE123468_12and16, file = "KEGG.UP_GSE123468_groupFam12&16.csv", row.names = FALSE)
save(kegg.up.GSE123468_12and16, file = "KEGG.UP_GSE123468_groupFam12&16.RData")
```

```{r}
kegg.dn.GSE123468_12and16 <- kegg.dn
kegg.dn.GSE123468_12and16 <- kegg.dn.GSE123468_12and16[order(kegg.dn.GSE123468_12and16$P.Down),]
write.csv(kegg.dn.GSE123468_12and16, file = "KEGG.DN_GSE123468_groupFam12&16.csv", row.names = FALSE)
save(kegg.dn.GSE123468_12and16, file = "KEGG.DN_GSE123468_groupFam12&16.RData")
```


## 2. Enrichment Analysis celltype:group


### 2. celltype:group, Family 12

```{r}
GeneID.UP <- entrezIDs[(resgroup.celltype12$logFC > 0 & 
                        resgroup.celltype12$adj.P.Val < 0.05)]
GeneID.DN <- entrezIDs[(resgroup.celltype12$logFC < 0 & 
                        resgroup.celltype12$adj.P.Val < 0.05)]
```


#### 2. celltype:group, Family 12: Calculation of enrichment for GO

```{r}
library(GO.db)
go <- goana(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:20,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:20,]
```


### Export of Results: GO, 2. celltype:group, Family 12

```{r}
go.up.GSE123468_12c <- go.up
go.up.GSE123468_12c <- go.up.GSE123468_12c[order(go.up.GSE123468_12c$P.Up),]
write.csv(go.up.GSE123468_12c, file = "GO.UP_GSE123468_group&celltype_Fam12.csv", row.names = FALSE)
save(go.up.GSE123468_12c, file = "GO.UP_GSE123468_group&celltype_Fam12.RData")
```

```{r}
go.dn.GSE123468_12c <- go.dn
go.dn.GSE123468_12c <- go.dn.GSE123468_12c[order(go.dn.GSE123468_12c$P.Down),]
write.csv(go.dn.GSE123468_12c, file = "GO.DN_GSE123468_group&celltype_Fam12.csv", row.names = FALSE)
save(go.dn.GSE123468_12c, file = "GO.DN_GSE123468_group&celltype_Fam12.RData")
```


#### 2. celltype:group, Family 12: Calculation of enrichment for KEGG

```{r}
kg <- kegga(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
kegg.up <- topKEGG(kg, sort = "up", number = Inf)
kegg.up[1:10,]
kegg.dn <- topKEGG(kg, sort = "down", number = Inf)
kegg.dn[1:10,]
```


### Export of Results: KEGG, 2. group:celltype, Family 12

```{r}
kegg.up.GSE123468_12c <- kegg.up
kegg.up.GSE123468_12c <- kegg.up.GSE123468_12c[order(kegg.up.GSE123468_12c$P.Up),]
write.csv(kegg.up.GSE123468_12c, file = "KEGG.UP_GSE123468_group&celltype_Fam12.csv", row.names = FALSE)
save(kegg.up.GSE123468_12c, file = "KEGG.UP_GSE123468_group&celltype_Fam12.RData")
```

```{r}
kegg.dn.GSE123468_12c <- kegg.dn
kegg.dn.GSE123468_12c <- kegg.dn.GSE123468_12c[order(kegg.dn.GSE123468_12c$P.Down),]
write.csv(kegg.dn.GSE123468_12c, file = "KEGG.DN_GSE123468_group&celltype_Fam12.csv", row.names = FALSE)
save(kegg.dn.GSE123468_12c, file = "KEGG.DN_GSE123468_group&celltype_Fam12.RData")
```


### 2. celltype:group, Family 16

```{r}
GeneID.UP <- entrezIDs[(resgroup.celltype16$logFC > 0 & 
                        resgroup.celltype16$adj.P.Val < 0.05)]
GeneID.DN <- entrezIDs[(resgroup.celltype16$logFC < 0 & 
                        resgroup.celltype16$adj.P.Val < 0.05)]
```


#### 2. celltype:group, Family 16: Calculation of enrichment for GO

```{r}
library(GO.db)
go <- goana(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:20,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:20,]
```


### Export of Results: GO,  2. group:celltype, Family 16

```{r}
go.up.GSE123468_16c <- go.up
go.up.GSE123468_16c <- go.up.GSE123468_16c[order(go.up.GSE123468_16c$P.Up),]
write.csv(go.up.GSE123468_16c, file = "GO.UP_GSE123468_group&celltype_Fam16.csv", row.names = FALSE)
save(go.up.GSE123468_16c, file = "GO.UP_GSE123468_group&celltype_Fam16.RData")
```

```{r}
go.dn.GSE123468_16c <- go.dn
go.dn.GSE123468_16c <- go.dn.GSE123468_16c[order(go.dn.GSE123468_16c$P.Down),]
write.csv(go.dn.GSE123468_16c, file = "GO.DN_GSE123468_group&celltype_Fam16.csv", row.names = FALSE)
save(go.dn.GSE123468_16c, file = "GO.DN_GSE123468_group&celltype_Fam16.RData")
```


#### 2. celltype:group, Family 16. Calculation of enrichment for KEGG

```{r}
kg <- kegga(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
kegg.up <- topKEGG(kg, sort = "up", number = Inf)
kegg.up[1:10,]
kegg.dn <- topKEGG(kg, sort = "down", number = Inf)
kegg.dn[1:10,]
```


### Export of Results: KEGG, 2. celltype:group, Family 16

```{r}
kegg.up.GSE123468_16c <- kegg.up
kegg.up.GSE123468_16c <- kegg.up.GSE123468_16c[order(kegg.up.GSE123468_16c$P.Up),]
write.csv(kegg.up.GSE123468_16c, file = "KEGG.UP_GSE123468_group&celltype_Fam16.csv", row.names = FALSE)
save(kegg.up.GSE123468_16c, file = "KEGG.UP_GSE123468_group&celltype_Fam16.RData")
```

```{r}
kegg.dn.GSE123468_16c <- kegg.dn
kegg.dn.GSE123468_16c <- kegg.dn.GSE123468_16c[order(kegg.dn.GSE123468_16c$P.Down),]
write.csv(kegg.dn.GSE123468_16c, file = "KEGG.DN_GSE123468_group&celltype_Fam16.csv", row.names = FALSE)
save(kegg.dn.GSE123468_16c, file = "KEGG.DN_GSE123468_group&celltype_Fam16.RData")
```


### 2. celltype:group: Families 12 & 16

```{r}
GeneID.UP.12 <- entrezIDs[(resgroup.celltype12$logFC > 0 & 
                        resgroup.celltype12$adj.P.Val < 0.05)]
GeneID.UP.16 <- entrezIDs[(resgroup.celltype16$logFC > 0 & 
                        resgroup.celltype16$adj.P.Val < 0.05)]
GeneID.UP <- union(GeneID.UP.12, GeneID.UP.16)

GeneID.DN.12 <- entrezIDs[(resgroup.celltype12$logFC < 0 & 
                        resgroup.celltype12$adj.P.Val < 0.05)]
GeneID.DN.16 <- entrezIDs[(resgroup.celltype16$logFC < 0 & 
                        resgroup.celltype16$adj.P.Val < 0.05)]
GeneID.DN <- union(GeneID.DN.12, GeneID.DN.16)
```


#### 2. celltype:group: Families 12 & 16: Calculation of enrichment for GO

```{r}
library(GO.db)
go <- goana(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:50,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:50,]
```


### Export of Results: GO,  2. group:celltype, Families 12 & 16

```{r}
go.up.GSE123468_12and16c <- go.up
go.up.GSE123468_12and16c <- go.up.GSE123468_12and16c[order(go.up.GSE123468_12and16c$P.Up),]
write.csv(go.up.GSE123468_12and16c, file = "GO.UP_GSE123468_group&celltype_Fam12&16.csv", row.names = FALSE)
save(go.up.GSE123468_12and16c, file = "GO.UP_GSE123468_group&celltype_Fam12&16.RData")
```

```{r}
go.dn.GSE123468_12and16c <- go.dn
go.dn.GSE123468_12and16c <- go.dn.GSE123468_12and16c[order(go.dn.GSE123468_12and16c$P.Down),]
write.csv(go.dn.GSE123468_12and16c, file = "GO.DN_GSE123468_group&celltype_Fam12&16.csv", row.names = FALSE)
save(go.dn.GSE123468_12and16c, file = "GO.DN_GSE123468_group&celltype_Fam12&16.RData")
```


#### 2. celltype:group: Families 12 & 16: Calculation of enrichment for KEGG

```{r}
kg <- kegga(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
kegg.up <- topKEGG(kg, sort = "up", number = Inf)
kegg.up[1:10,]
kegg.dn <- topKEGG(kg, sort = "down", number = Inf)
kegg.dn[1:10,]
```


### Export of Results: KEGG, 2. group:celltype, Families 12 & 16

```{r}
kegg.up.GSE123468_12and16c <- kegg.up
kegg.up.GSE123468_12and16c <- kegg.up.GSE123468_12and16c[order(kegg.up.GSE123468_12and16c$P.Up),]
write.csv(kegg.up.GSE123468_12and16c, file = "KEGG.UP_GSE123468_group&celltype_Fam12&16.csv", row.names = FALSE)
save(kegg.up.GSE123468_12and16c, file = "KEGG.UP_GSE123468_group&celltype_Fam12&16.RData")
```

```{r}
kegg.dn.GSE123468_12and16c <- kegg.dn
kegg.dn.GSE123468_12and16c <- kegg.dn.GSE123468_12and16c[order(kegg.dn.GSE123468_12and16c$P.Down),]
write.csv(kegg.dn.GSE123468_12and16c, file = "KEGG.DN_GSE123468_group&celltype_Fam12&16.csv", row.names = FALSE)
save(kegg.dn.GSE123468_12and16c, file = "KEGG.DN_GSE123468_group&celltype_Fam12&16.RData")
```


# Used software

```{r}
sessionInfo()
```
