---
title: "Analysis of GSE15090"
author: "T. SchĂ¤tzl and M. Kohl"
date: "May 2022"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
## setup fĂĽr die Chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


# Import of Data

The data are included in dataset GSE15090 of Gene Expression Omnibus:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE36398

The tar-file was unzipped into a subfolder called GSE36398_RAW. 
We removed all files containing data about asymptotic people.

We import the data from the CEL-files using Bioconductor package oligo.

```{r}
library(affy)
rawData <- ReadAffy(celfile.path = "./GSE15090_RAW")
rawData
```


# Quality Control

```{r, eval=FALSE}
library(arrayQualityMetrics)
arrayQualityMetrics(rawData, outdir = "QC", force = TRUE, do.logtransform = TRUE)
```

```{r}
library(MKomics)
COR <- cor(exprs(rawData), method = "spearman")
corPlot2(COR, minCor = min(COR), fontsize.axis = 8, 
         title = "Spearman Correlation of Raw Data")
```


# Preprocessing

We use RMA to preprocess the data.

```{r, cache = TRUE}
normData <- rma(rawData)
normData
```


# Quality Control of Preprocessed Data

```{r}
library(vsn)
meanSdPlot(normData)
```

```{r}
## range = 0: whiskers till most extreme data points
boxplot(exprs(normData), range = 0, main = "RMA", las = 2, 
        cex.axis = 0.75)
```

```{r}
COR.rma <- cor(exprs(normData), method = "spearman")
corPlot2(COR.rma, minCor = min(COR.rma), fontsize.axis = 8, 
         title = "Spearman Correlation of Preprocessed Data")
```

```{r}
SampleInfo <- read.csv(file = "GSE15090_SampleInfo.csv", 
                       na.strings = "N/A")
SampleInfo
```

```{r}
ExportData <- t(exprs(normData))
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE15090data.rds")
```


# Statistical Analysis

## Gene list

```{r}
library(org.Hs.eg.db)
library(hgu133plus2.db)
GeneID <- select(hgu133plus2.db, keys = keys(hgu133plus2.db), 
                 columns = c("ENSEMBL", "ENTREZID", "SYMBOL"))
```

```{r}
library(openxlsx)
GeneSel <- read.xlsx("FSHDlist.xlsx")
GeneSel$ENSEMBL <- sapply(strsplit(GeneSel$gene_id, "\\."), "[", 1)
dim(GeneSel)
```

```{r}
sum(GeneSel$ENSEMBL %in% GeneID$ENSEMBL)
GeneSel[!(GeneSel$ENSEMBL %in% GeneID$ENSEMBL),]
```

```{r}
GeneID.GeneSel <- GeneID[which(GeneID$ENSEMBL %in% GeneSel$ENSEMBL),]
GeneID.GeneSel <- GeneID.GeneSel[!duplicated(GeneID.GeneSel),]
GeneID.GeneSel
```

```{r}
ProbeID.GeneSel <- unique(GeneID.GeneSel$PROBEID)
sel.GeneSel <- rownames(exprs(normData)) %in% ProbeID.GeneSel
normData.GeneSel <- exprs(normData)[sel.GeneSel,]
```

```{r}
res.GeneSel <- mod.t.test(x = normData.GeneSel, group = SampleInfo$group)
colSums(res.GeneSel[,c(5,6)] < 0.05)
# results: 8 x p.val & 0 x adj.p.val 

sig.p <- row.names(res.GeneSel)[which(res.GeneSel[,5]<0.05)]
sig.p
GeneID.sig.p <- GeneID.GeneSel[GeneID.GeneSel$PROBEID %in% sig.p,]
GeneID.sig.p
unique(GeneID.sig.p$SYMBOL)
```


### Export of Results, Gene list

```{r}
res.GeneSel.GSE15090 <- as.data.frame(res.GeneSel)
res.GeneSel.GSE15090$PROBEID <- ProbeID.GeneSel
res.GeneSel.GSE15090 <- merge(res.GeneSel.GSE15090, GeneID.GeneSel, by = "PROBEID", all.x = TRUE)
res.GeneSel.GSE15090 <- res.GeneSel.GSE15090[order(res.GeneSel.GSE15090$p.value),]
write.csv(res.GeneSel.GSE15090, file = "ResultsGeneSel_GSE15090.csv", row.names = FALSE)
save(res.GeneSel.GSE15090, file = "ResultsGeneSel_GSE15090.RData")
```


## All genes

### mod. t-test

```{r}
library(MKomics)
res <- mod.t.test (x = exprs(normData), group = SampleInfo$group)
colSums(res[,c(5,6)] < 0.05)
# no results for adj. p value < 5%
```


# Used software

```{r}
sessionInfo()
```

