---
title: "Analysis of GSE115650"
author: "T. Schaetzl and M. Kohl"
date: "August 2022"
output:
  rmarkdown::html_document:
    theme: united
    highlight: tango
    toc: true
---

# Loading of required packages

```{r}
library(tximport)
library(MKomics)
library(vsn)
library(limma)
library(MKinfer)
library(VennDiagram)
library(edgeR)
library(GO.db)
library(org.Hs.eg.db)
```


# Import of Salmon Quantification data

```{r}
SampleInfo <- read.csv("GSE115650_SampleInfo.csv", na.strings = "N/A")
SampleInfo
txi <- readRDS(file = "SRP150311.rds")
head(txi$counts)
cts <- txi$counts
cts <- cts[,order(colnames(cts))]
#Removing mosaic sample
cts <- cts[,SampleInfo$subject != "01-0032"]
SampleInfo <- SampleInfo[SampleInfo$subject != "01-0032",]
sort(colSums(cts)/1e6)
```


# Quality control of raw data

```{r}
cts0 <- cts
cts0[cts0 == 0] <- min(cts0[cts0 != 0])
boxplot(cts0, log = "y", range = 0, las = 2)
```

```{r}
Cor <- cor(cts, method = "spearman")
corPlot2(Cor, minCor = min(Cor), 
         #labels = paste(SampleInfo$subject),
         title = "Spearman correlation")
```


# Preprocessing

```{r}
dge <- DGEList(counts=cts)
group <- SampleInfo$group
design <- model.matrix(~ group)
keep <- filterByExpr(dge, design)
dge <- dge[keep,,keep.lib.sizes=FALSE]
dge <- calcNormFactors(dge)
voomData <- voom(dge, design, plot=FALSE)
```


## Quality control of filtered data

```{r}
boxplot(voomData$E, range = 0, las = 2)
```

```{r}
Cor <- cor(voomData$E, method = "spearman")
corPlot2(Cor, minCor = min(Cor), 
         #labels=paste(SampleInfo$subject),
         title = "Spearman correlation")
```

```{r}
meanSdPlot(voomData$E)
```

```{r}
library(org.Hs.eg.db)
symbols <- sapply(strsplit(rownames(voomData$E), "\\|"), "[", 6)
ensemblIDs <- sapply(strsplit(rownames(voomData$E), "\\|"), "[", 2)
ensemblIDs <- sapply(strsplit(ensemblIDs, "\\."), "[", 1)
entrezIDs <- mapIds(org.Hs.eg.db, keys = ensemblIDs, column = "ENTREZID", 
                    keytype = "ENSEMBL", multiVals = "first")
```

```{r}
ExportData <- t(voomData$E) 
colnames(ExportData) <- ensemblIDs
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE115650data.rds")
```


# Statistical Analysis

## Gene list

```{r}
library(openxlsx)
GeneSel <- read.xlsx("FSHDlist.xlsx")
GeneSel$ENSEMBL <- sapply(strsplit(GeneSel$gene_id, "\\."), "[", 1)
dim(GeneSel)
```

```{r}
voomData.Genesel <- voomData[ensemblIDs %in% GeneSel$ENSEMBL,]
GeneSel[!(GeneSel$ENSEMBL %in% ensemblIDs),]
```

```{r}
library(org.Hs.eg.db)
symbols.Genesel <- sapply(strsplit(rownames(voomData.Genesel$E), "\\|"), "[", 6)
ensemblIDs.Genesel <- sapply(strsplit(rownames(voomData.Genesel$E), "\\|"), "[", 2)
ensemblIDs.Genesel <- sapply(strsplit(ensemblIDs.Genesel, "\\."), "[", 1)
entrezIDs.Genesel <- mapIds(org.Hs.eg.db, keys = ensemblIDs.Genesel, column = "ENTREZID", 
                    keytype = "ENSEMBL", multiVals = "first")
```

```{r}
library(MKomics)
fit <- lmFit(voomData.Genesel$E,design)
fit <- eBayes(fit)
res.GeneSel <- topTable(fit, number = Inf)
sum(res.GeneSel$P.Value < 0.05)
# 72 results
sort(unique(symbols.Genesel[res.GeneSel$P.Value < 0.05]))
sum(res.GeneSel$adj.P.Val < 0.05)
# 31 results
sort(unique(symbols.Genesel[res.GeneSel$adj.P.Val < 0.05]))
```


### Export of Results, Gene list

```{r}
res.GeneSel.GSE115650 <- as.data.frame(res.GeneSel)
res.GeneSel.GSE115650$ENSEMBLE <- ensemblIDs.Genesel
res.GeneSel.GSE115650 <- res.GeneSel.GSE115650[order(res.GeneSel.GSE115650$P.Val),]
write.csv(res.GeneSel.GSE115650, file = "ResultsGeneSel_GSE115650.csv", row.names = FALSE)
save(res.GeneSel.GSE115650, file = "ResultsGeneSel_GSE115650.RData")
```


## All genes

```{r}
library(MKomics)
fit <- lmFit(voomData$E,design)
fit <- eBayes(fit)
res <- topTable(fit, number = Inf)
sum(res$P.Value < 0.05)
sum(res$adj.P.Val < 0.05)
# 1175 results for adj. p value < 5%
sort(unique(symbols[res$adj.P.Val < 0.05]))
```


### Export of Results

```{r}
res.GSE115650 <- as.data.frame(res)
res.GSE115650$ENSEMBLE <- ensemblIDs
res.GSE115650 <- res.GSE115650[order(res.GSE115650$adj.P.Val),]
write.csv(res.GSE115650, file = "ResultsGSE115650.csv", row.names = FALSE)
save(res.GSE115650, file = "ResultsGSE115650.RData")
```


## Heatmap 

```{r}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
## Data selection
sel <-res$adj.P.Val < 0.05
selData <- voomData$E[sel,]
selData <- selData - rowMeans(selData, na.rm = TRUE)
colnames(selData) <- paste(SampleInfo$subject)
## Color definition
col1 <- rev(brewer.pal(n = 8, name = "RdYlBu"))
col2 <- brewer.pal(n = 3, name = "Set1")[c(3,1)]
col3 <- brewer.pal(n = 3, name = "Set2")[c(3,1)]
col4 <- brewer.pal(n = 3, name = "Dark2")[c(2,4)]
mycol <- colorRamp2(seq(-4, 4, length = 128), #16fache Expression
                    colorRampPalette(col1)(128)) 
## Annotation
df <- data.frame(group = SampleInfo$group)
ha <- HeatmapAnnotation(df = df,
                        col = list(group = c("control" = col2[1],
                                              "FSHD" = col2[2])))
## Drawing the heatmap
Heatmap(selData, col = mycol, name = "log2-FC", show_row_names = FALSE,
        use_raster = FALSE,
        top_annotation = ha, show_column_names = TRUE,
        column_names_gp = gpar(fontsize = 8),
        clustering_distance_columns = "euclidean",
        clustering_distance_rows = "euclidean",
        show_row_dend = TRUE,
        cluster_columns = TRUE, show_column_dend = TRUE,
        column_title = "p value < 0.05 and |log2-FC| > log2(1.5)",
        show_heatmap_legend = TRUE)
```


### Enrichment Analysis

```{r}

GeneID.UP <- entrezIDs[(res$logFC > 0 & 
                        res$adj.P.Val < 0.05)]
GeneID.DN <- entrezIDs[(res$logFC > 0 & 
                        res$adj.P.Val < 0.05)]
```


#### Calculation of enrichment for GO

```{r}
library(GO.db)
go <- goana(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:20,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:20,]
```


### Export of Results: GO

```{r}
go.up.GSE115650 <- go.up
go.up.GSE115650 <- go.up.GSE115650[order(go.up.GSE115650$P.Up),]
write.csv(go.up.GSE115650, file = "GO.UP_GSE115650.csv", row.names = FALSE)
save(go.up.GSE115650, file = "GO.UP_GSE115650.RData")
```

```{r}
go.dn.GSE115650 <- go.dn
go.dn.GSE115650 <- go.dn.GSE115650[order(go.dn.GSE115650$P.Down),]
write.csv(go.dn.GSE115650, file = "GO.DN_GSE115650.csv", row.names = FALSE)
save(go.dn.GSE115650, file = "GO.DN_GSE115650.RData")
```


#### Calculation of enrichment for KEGG

```{r}
kg <- kegga(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
kegg.up <- topKEGG(kg, sort = "up", number = Inf)
kegg.up[1:20,]
kegg.dn <- topKEGG(kg, sort = "down", number = Inf)
kegg.dn[1:20,]
```


### Export of Results: KEGG

```{r}
kegg.up.GSE115650 <- kegg.up
kegg.up.GSE115650 <- kegg.up.GSE115650[order(kegg.up.GSE115650$P.Up),]
write.csv(kegg.up.GSE115650, file = "KEGG.UP_GSE115650.csv", row.names = FALSE)
save(kegg.up.GSE115650, file = "KEGG.UP_GSE115650.RData")
```

```{r}
kegg.dn.GSE115650 <- kegg.dn
kegg.dn.GSE115650 <- kegg.dn.GSE115650[order(kegg.dn.GSE115650$P.Down),]
write.csv(kegg.dn.GSE115650, file = "KEGG.DN_GSE115650.csv", row.names = FALSE)
save(kegg.dn.GSE115650, file = "KEGG.DN_GSE115650.RData")
```


# Used software

```{r}
sessionInfo()
```
