---
title: "Analysis of GSE26852"
author: "T. Schätzl and M. Kohl"
date: "May 2022"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
---


```{r setup, include=FALSE}
## setup für die Chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


# Import of Data

The data are included in dataset GSE26852 of Gene Expression Omnibus:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26852

The tar-file was unzipped into a subfolder called GSE26852_RAW. 
We removed all files containing data about inflammatory myopathies (n=7) and dysferlinopathies (n=4).

We import the data from the probe profiles using Bioconductor package lumi.

```{r}
library(lumi)
Ilu.Daten <- lumiR(fileName = "GSE26852_non-normalized.txt")
Ilu.Daten <- Ilu.Daten[,1:15]
Ilu.Daten
SampleInfo <- read.csv(file = "GSE26852_SampleInfo.csv", na.strings = "N/A")
SampleInfo
Ilu.Daten <- Ilu.Daten[,SampleInfo$disease_state != "T2-STIR-"]
SampleInfo <- SampleInfo[SampleInfo$disease_state != "T2-STIR-",]
ProbeID <- read.delim(file = "GSE26852_non-normalized.txt", skip = 1)$X
library(beadarray)
Ilu.Daten2 <- readBeadSummaryData(dataFile = "GSE26852_non-normalized.txt", 
                                  skip = 1, ProbeID = "ProbeID")
Ilu.Daten2 <- Ilu.Daten2[,1:15]
Ilu.Daten2
Ilu.Daten2 <- Ilu.Daten[,SampleInfo$disease_state != "T2-STIR-"]
SampleInfo <- SampleInfo[SampleInfo$disease_state != "T2-STIR-",]

```


# Quality Control

```{r}
library(arrayQualityMetrics)
arrayQualityMetrics(Ilu.Daten2, outdir = "QC", force = TRUE,
                    do.logtransform = TRUE)
```


# Similarity between arrays

```{r}
library(MKomics)
COR <- cor(exprs(Ilu.Daten), method = "spearman")
corPlot2(COR, minCor = min(COR), labels = pData(Ilu.Daten)[,1], 
        title = "Spearman Correlation")
```


# Boxplots of expression values

```{r}
lumi::boxplot(Ilu.Daten)
```


# Preprocessing

```{r}
Ilu.rsn <- rsn(lumiT(Ilu.Daten, method = "log2"))
```


# Quality Control of Preprocessed Data

```{r}
library(vsn)
meanSdPlot(Ilu.rsn)
```


# Boxplots of normalized values

```{r}
boxplot(Ilu.rsn, main = "rsn")
```


# Similarity between arrays

```{r}
COR.rsn <- cor(exprs(Ilu.rsn))
corPlot2(COR.rsn, minCor = min(COR), title = "rsn")
```

```{r}
ExportData <- t(exprs(Ilu.rsn))
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE26852data.rds")
```


## Gene list

```{r}
library(org.Hs.eg.db)
library(illuminaHumanv3.db)
GeneID <- select(illuminaHumanv3.db, keys = keys(illuminaHumanv3.db),
                 columns = c("ENSEMBL", "ENTREZID", "SYMBOL"))
```

```{r}
library(openxlsx)
GeneSel <- read.xlsx("FSHDlist.xlsx")
GeneSel$ENSEMBL <- sapply(strsplit(GeneSel$gene_id, "\\."), "[", 1)
dim(GeneSel)
```

```{r}
sum(GeneSel$ENSEMBL %in% GeneID$ENSEMBL)
GeneSel[!(GeneSel$ENSEMBL %in% GeneID$ENSEMBL),]
```

```{r}
GeneID.GeneSel <- GeneID[which(GeneID$ENSEMBL %in% GeneSel$ENSEMBL),]
GeneID.GeneSel <- GeneID.GeneSel[!duplicated(GeneID.GeneSel),]
GeneID.GeneSel
```

```{r}
ProbeID.GeneSel <- unique(GeneID.GeneSel$PROBEID)
sel.GeneSel <- ProbeID %in% ProbeID.GeneSel
Ilu.rsn.GeneSel <- exprs(Ilu.rsn)[sel.GeneSel,]
```

```{r}
library(MKomics)
res.GeneSel <- mod.t.test(x = Ilu.rsn.GeneSel, group = SampleInfo$group)
res.GeneSel
colSums(res.GeneSel[,c(5,6)] < 0.05)
# results: 48 x p.val & 15 x adj.p.val
row.names(res.GeneSel)[which(res.GeneSel[,6]<0.05)]
row.names(res.GeneSel)[which(res.GeneSel[,5]<0.05)]
```


### Export of Results, Gene list:

```{r}
res.GeneSel.GSE26852_stir <- as.data.frame(res.GeneSel)
res.GeneSel.GSE26852_stir$PROBEID <- ProbeID.GeneSel
res.GeneSel.GSE26852_stir <- merge(res.GeneSel.GSE26852_stir, GeneID, by = "PROBEID", all.x = TRUE)
res.GeneSel.GSE26852_stir <- res.GeneSel.GSE26852_stir[order(res.GeneSel.GSE26852_stir$p.value),]
write.csv(res.GeneSel.GSE26852_stir, file = "ResultsGeneSel_GSE26852.csv", row.names = FALSE)
save(res.GeneSel.GSE26852_stir, file = "ResultsGeneSel_GSE26852.RData")
```


## All genes

### mod. t-test

```{r}
library(MKomics)
res <- mod.t.test(x = exprs(Ilu.rsn), group = SampleInfo$group)
colSums(res[,c(5,6)] < 0.05)
# 3489 results for adj. p value < 5%
```


### Export of Results

```{r}
res.GSE26852_stir <- as.data.frame(res)
res.GSE26852_stir$PROBEID <- ProbeID
res.GSE26852_stir <- merge(res.GSE26852_stir, GeneID, by = "PROBEID", all.x = TRUE)
res.GSE26852_stir <- res.GSE26852_stir[order(res.GSE26852_stir$adj.p.value),]
write.csv(res.GSE26852_stir, file = "ResultsGSE26852.csv", row.names = FALSE)
save(res.GSE26852_stir, file = "ResultsGSE26852.RData")
```


### Heatmap 

```{r}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
## Data selection
sel <- res$p.value < 0.05 & abs(res$`difference in means`) > 1
selData <- exprs(Ilu.rsn)[sel,]
selData <- selData - rowMeans(selData, na.rm = TRUE)
colnames(selData) <- as.character(SampleInfo$group)
## Setting colors
col1 <- rev(brewer.pal(n = 8, name = "RdYlBu"))
col2 <- brewer.pal(n = 3, name = "Set1")[c(3,1)]
mycol <- colorRamp2(seq(-2, 2, length = 128), colorRampPalette(col1)(128))
## Annotation
df <- data.frame(group = SampleInfo$group)
ha <- HeatmapAnnotation(df = df,
                        col = list(Gruppe = c("control" = col2[1],
                                             "FSHD" = col2[2])))
## Drawing the Heatmap
Heatmap(selData, col = mycol, name = "log2-FC", show_row_names = FALSE,
        top_annotation = ha, show_column_names = TRUE,
        column_names_gp = gpar(fontsize = 8),
        clustering_distance_columns = "pearson",
        show_row_dend = FALSE, km = 4,
        cluster_columns = TRUE, show_column_dend = TRUE,
        column_title = "Adj. p Wert < 0.05 und |log2-FC| > 1",
        show_heatmap_legend = TRUE)
```


### Enrichment Analysis

```{r}
library(org.Hs.eg.db)
library(illuminaHumanv3.db)
GeneID <- select(illuminaHumanv3.db, keys = keys(illuminaHumanv3.db), 
                 columns = "ENTREZID")
keys.UP <- ProbeID[(res$`difference in means` > 0 & 
                                   res$p.value < 0.05)]
keys.DN <- ProbeID[(res$`difference in means` < 0 & 
                                   res$p.value < 0.05)]
GeneID.UP <- select(illuminaHumanv3.db, keys = keys.UP, columns = "ENTREZID")
GeneID.DN <- select(illuminaHumanv3.db, keys = keys.DN, columns = "ENTREZID")
```


#### Calculation of enrichment for GO

```{r}
library(limma)
go <- goana(list(Up = unique(GeneID.UP$ENTREZID), 
                 Down = unique(GeneID.DN$ENTREZID)), 
            universe = unique(GeneID$ENTREZID), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:40,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:40,]
```


### Export of Results: GO

```{r}
go.up.GSE26852_stir <- go.up
go.up.GSE26852_stir <- go.up.GSE26852_stir[order(go.up.GSE26852_stir$P.Up),]
write.csv(go.up.GSE26852_stir, file = "GO.UP_GSE26852.csv", row.names = FALSE)
save(go.up.GSE26852_stir, file = "GO.UP_GSE26852.RData")
```

```{r}
go.dn.GSE26852_stir <- go.dn
go.dn.GSE26852_stir <- go.dn.GSE26852_stir[order(go.dn.GSE26852_stir$P.Down),]
write.csv(go.dn.GSE26852_stir, file = "GO.DN_GSE26852.csv", row.names = FALSE)
save(go.dn.GSE26852_stir, file = "GO.DN_GSE26852.RData")
```


#### Calculation of enrichment for KEGG

```{r}
kg <- kegga(list(Up = unique(GeneID.UP$ENTREZID), 
                 Down = unique(GeneID.DN$ENTREZID)), 
            universe = unique(GeneID$ENTREZID), 
            species = "Hs")
kegg.up <- topKEGG(kg, sort = "up", number = Inf)
kegg.up[1:20,]
kegg.dn <- topKEGG(kg, sort = "down", number = Inf)
kegg.dn[1:20,]
```


### Export of Results: KEGG

```{r}
kegg.up.GSE26852_stir <- kegg.up
kegg.up.GSE26852_stir <- kegg.up.GSE26852_stir[order(kegg.up.GSE26852_stir$P.Up),]
write.csv(kegg.up.GSE26852_stir, file = "KEGG.UP_GSE26852.csv", row.names = FALSE)
save(kegg.up.GSE26852_stir, file = "KEGG.UP_GSE26852.RData")
```

```{r}
kegg.dn.GSE26852_stir <- kegg.dn
kegg.dn.GSE26852_stir <- kegg.dn.GSE26852_stir[order(kegg.dn.GSE26852_stir$P.Down),]
write.csv(kegg.dn.GSE26852_stir, file = "KEGG.DN_GSE26852.csv", row.names = FALSE)
save(kegg.dn.GSE26852_stir, file = "KEGG.DN_GSE26852.RData")
```


# Used software

```{r}
sessionInfo()
```

