---
title: "Analysis of GSE26145"
author: "T. Schätzl and M. Kohl"
date: "May 2022"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
---


```{r setup, include=FALSE}
## setup für die Chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


# Import of Data

The data are included in dataset GSE26145 of Gene Expression Omnibus:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26145

The tar-file was unzipped into a subfolder called GSE26145_RAW. 
We removed the disease control.

We import the data from the CEL-files using Bioconductor package oligo.

```{r}
library(oligo)
celFiles <- list.files("./GSE26145_RAW", full.names = TRUE)
rawData <- read.celfiles(celFiles)
rawData
```


# Quality Control

```{r, eval=FALSE}
library(arrayQualityMetrics)
arrayQualityMetrics(rawData, outdir = "QC", force = TRUE, do.logtransform = TRUE)
```

In addition, we take a look at RLE (Relative Log Expression) and NUSE (Normalized Unscaled Standard Errors)

```{r}
fit1 <- fitProbeLevelModel(rawData)
op <- par(mar = c(12,4,4,2)+0.1)
RLE(fit1, las = 2, cex.axis = 0.75)
NUSE(fit1, las = 2, cex.axis = 0.75)
par(op)
```

```{r}
library(MKomics)
COR <- cor(exprs(rawData), method = "spearman")
corPlot2(COR, minCor = min(COR), fontsize.axis = 8, 
         title = "Spearman Correlation of Raw Data")
```


# Preprocessing

We use RMA to preprocess the data.

```{r, cache = TRUE}
normData <- rma(rawData)
normData
```


# Quality Control of Preprocessed Data

```{r}
library(vsn)
meanSdPlot(normData)
```

```{r}
## range = 0: whiskers till most extreme data points
boxplot(exprs(normData), range = 0, main = "RMA", las = 2, 
        cex.axis = 0.75)
```

```{r}
COR.rma <- cor(exprs(normData), method = "spearman")
corPlot2(COR.rma, minCor = min(COR.rma), fontsize.axis = 8, 
         title = "Spearman Correlation of Preprocessed Data")
```

Batch effect scan date?

```{r}
SampleInfo <- read.csv(file = "GSE26145_SampleInfo.csv", na.strings = "N/A")
SampleInfo
```


# Removing of Batch Effects

The batch effects cannot be removed because we do not know from the experiment design whether it is a batch effect or an effect between the groups FSHD vs. controls

```{r}
ExportData <- t(exprs(normData))
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE26145data.rds")
```


# Statistical Analysis

## Gene list

```{r}
library(org.Hs.eg.db)
library(huex10sttranscriptcluster.db)
GeneID <- select(huex10sttranscriptcluster.db, keys = keys(huex10sttranscriptcluster.db), 
                 columns = c("ENSEMBL", "ENTREZID", "SYMBOL"))
```

```{r}
library(openxlsx)
GeneSel <- read.xlsx("FSHDlist.xlsx")
GeneSel$ENSEMBL <- sapply(strsplit(GeneSel$gene_id, "\\."), "[", 1)
dim(GeneSel)
```

```{r}
sum(GeneSel$ENSEMBL %in% GeneID$ENSEMBL)
GeneSel[!(GeneSel$ENSEMBL %in% GeneID$ENSEMBL),]
```

```{r}
GeneID.GeneSel <- GeneID[which(GeneID$ENSEMBL %in% GeneSel$ENSEMBL),]
GeneID.GeneSel <- GeneID.GeneSel[!duplicated(GeneID.GeneSel),]
GeneID.GeneSel
```

```{r}
ProbeID.GeneSel <- unique(GeneID.GeneSel$PROBEID)
sel.GeneSel <- rownames(exprs(normData)) %in% ProbeID.GeneSel
normData.GeneSel <- exprs(normData)[sel.GeneSel,]
```

```{r}
library(limma)
subject <- factor(SampleInfo$subject)
cell_type <- factor(SampleInfo$cell_type)
group <- factor(SampleInfo$group)
design <- model.matrix(~cell_type*group)
fit <- lmFit(normData.GeneSel, design)
fit <- eBayes(fit)

resgroup <- topTable(fit, coef="groupFSHD", number = Inf)
colSums(resgroup[,c(4,5)] < 0.05) 
# results: 2 x p.val & 0 x adj.p.val
sig.p <- row.names(resgroup)[which(resgroup[,4]<0.05)]
sig.p
GeneID.sig.p <- GeneID.GeneSel[GeneID.GeneSel$PROBEID %in% sig.p,]
GeneID.sig.p
unique(GeneID.sig.p$SYMBOL)

rescelltype <- topTable(fit, coef="cell_typeMT", number = Inf)
colSums(rescelltype[,c(4,5)] < 0.05)
# results: 2 x p.val & 0 x adj.p.val
sig.p <- row.names(rescelltype)[which(rescelltype[,4]<0.05)]
sig.p
GeneID.sig.p <- GeneID.GeneSel[GeneID.GeneSel$PROBEID %in% sig.p,]
GeneID.sig.p
unique(GeneID.sig.p$SYMBOL)

rescelltypegroup <- topTable(fit, coef="cell_typeMT:groupFSHD", number = Inf)
colSums(rescelltypegroup[,c(4,5)] < 0.05)
# no results
```


## Export of Results, Gene list

```{r}
temp1 <- resgroup
colnames(temp1) <- paste0("group: ", colnames(resgroup))
temp2 <- rescelltype
colnames(temp2) <- paste0("celltype: ", colnames(rescelltype))
temp3 <- rescelltypegroup
colnames(temp3) <- paste0("group:celltype: ", 
                                      colnames(rescelltypegroup))
res.GeneSel.GSE26145 <- cbind(temp1[order(rownames(temp1)),], temp2[order(rownames(temp2)),], temp3[order(rownames(temp3)),])
res.GeneSel.GSE26145$PROBEID <- sort(rownames(temp1))
res.GeneID.GeneSel<-GeneID.GeneSel[GeneID.GeneSel$PROBEID %in% rownames(temp1),]
res.GeneSel.GSE26145 <- merge(res.GeneSel.GSE26145,res.GeneID.GeneSel, by="PROBEID", all=TRUE)
rm(temp1,temp2, temp3)
write.csv(res.GeneSel.GSE26145, file = "ResultsGeneSel_GSE26145.csv")
save(res.GeneSel.GSE26145, file = "ResultsGeneSel_GSE26145.RData")
```


## All genes

### Analysis of variance

```{r}
library(limma)
subject <- factor(SampleInfo$subject)
cell_type <- factor(SampleInfo$cell_type)
group <- factor(SampleInfo$group)
design <- model.matrix(~cell_type*group)
fit <- lmFit(exprs(normData), design)
fit <- eBayes(fit)

resgroup <- topTable(fit, coef="groupFSHD", number = Inf)
sum(resgroup$P.Value < 0.05)
sum(resgroup$adj.P.Value < 0.05)
# no results for adj. p value < 5%

rescelltype <- topTable(fit, coef="cell_typeMT", number = Inf)
sum(rescelltype$P.Val < 0.05)
sum(rescelltype$adj.P.Val < 0.05)
# no results for adj. p value < 5%

rescelltypegroup <- topTable(fit, coef="cell_typeMT:groupFSHD", number = Inf)
sum(rescelltypegroup$P.Val < 0.05)
sum(rescelltypegroup$adj.P.Val < 0.05)
# no results for adj. p value < 5%
```


# Used Software

```{r}
sessionInfo()
```

