---
title: "Analysis of GSE36398"
author: "T. Schätzl and M. Kohl"
date: "May 2022"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
---


```{r setup, include=FALSE}
## setup für die Chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


# Import of Data

The data are included in dataset GSE36398 of Gene Expression Omnibus:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE36398

The tar-file was unzipped into a subfolder called GSE36398_RAW. 
We removed all files with deltoid muscle data because there are both biceps and deltoid muscle biopsies on the same individual with an associated family control in the present data set. This is because the biceps is more affected compared with the deltoid muscle in FSHD. 

We import the data from the CEL-files using Bioconductor package oligo.

```{r}
library(oligo)
celFiles <- list.files("./GSE36398_RAW", full.names = TRUE)
rawData <- read.celfiles(celFiles)
rawData
```


# Quality Control

```{r, eval=FALSE}
library(arrayQualityMetrics)
arrayQualityMetrics(rawData, outdir = "QC", force = TRUE, do.logtransform = TRUE)
```

In addition, we take a look at RLE (Relative Log Expression) and NUSE (Normalized Unscaled Standard Errors)

```{r}
fit1 <- fitProbeLevelModel(rawData)
op <- par(mar = c(12,4,4,2)+0.1)
RLE(fit1, las = 2, cex.axis = 0.75)
NUSE(fit1, las = 2, cex.axis = 0.75)
par(op)
```

```{r}
library(MKomics)
COR <- cor(exprs(rawData), method = "spearman")
corPlot2(COR, minCor = min(COR), fontsize.axis = 8, 
         title = "Spearman Correlation of Raw Data")
```


# Preprocessing

We use RMA to preprocess the data.

```{r, cache = TRUE}
normData <- rma(rawData)
normData
```


# Quality Control of Preprocessed Data

```{r}
library(vsn)
meanSdPlot(normData)
```

```{r}
## range = 0: whiskers till most extreme data points
boxplot(exprs(normData), range = 0, main = "RMA", las = 2, 
        cex.axis = 0.75)
```

```{r}
COR.rma <- cor(exprs(normData), method = "spearman")
corPlot2(COR.rma, minCor = min(COR.rma), fontsize.axis = 8, signifBar = 3,
         title = "Spearman Correlation of Preprocessed Data")
```

Batch effect scan date!

```{r}
SampleInfo <- read.csv(file = "GSE36398_SampleInfo.csv")
SampleInfo
```


# Removing of Batch Effects 

We use the empirical Bayes approach proposed by W.E. Johnson, C. Li, and A. Rabinovic. Adjusting batch effects in microarray data using empirical bayes methods. Biostatistics,8(1):118–127, 2007.

```{r}
library(sva)
scanDate <- sapply(strsplit(SampleInfo$scan_date, " |T"), "[", 1)
DF <- data.frame(group = as.factor(SampleInfo$group), 
                 scanDate = scanDate)
mod.batch <- model.matrix(~ group, data=DF)
normDataBatch <- ComBat(dat = normData, 
                        batch = scanDate, mod = mod.batch, 
                        par.prior = TRUE, mean.only = TRUE)
```

```{r}
meanSdPlot(normDataBatch)
```

```{r}
## range = 0: whiskers till most extreme data points
boxplot(normDataBatch, range = 0, main = "Parametric ComBat", las = 2, 
        cex.axis = 0.75)
```

```{r}
COR.param <- cor(normDataBatch, method = "spearman")
corPlot2(COR.param, minCor = min(COR.param), fontsize.axis = 8, 
         title = "Spearman Correlation of Batch Corrected Data", 
         signifBar = 3)
```

```{r}
ExportData <- t(normDataBatch)
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE36398data.rds")
```


# Statistical Analysis

## Gene list

```{r}
library(org.Hs.eg.db)
library(hugene10sttranscriptcluster.db)
GeneID <- select(hugene10sttranscriptcluster.db, keys = keys(hugene10sttranscriptcluster.db), 
                 columns = c("ENSEMBL", "ENTREZID", "SYMBOL"))
```

```{r}
library(openxlsx)
GeneSel <- read.xlsx("FSHDlist.xlsx")
GeneSel$ENSEMBL <- sapply(strsplit(GeneSel$gene_id, "\\."), "[", 1)
dim(GeneSel)
```

```{r}
sum(GeneSel$ENSEMBL %in% GeneID$ENSEMBL)
GeneSel[!(GeneSel$ENSEMBL %in% GeneID$ENSEMBL),]
```

```{r}
GeneID.GeneSel <- GeneID[which(GeneID$ENSEMBL %in% GeneSel$ENSEMBL),]
GeneID.GeneSel <- GeneID.GeneSel[!duplicated(GeneID.GeneSel),]
GeneID.GeneSel
```

```{r}
ProbeID.GeneSel <- unique(GeneID.GeneSel$PROBEID)
sel.GeneSel <- rownames(exprs(normData)) %in% ProbeID.GeneSel
normDataBatch.GeneSel <- normDataBatch[sel.GeneSel,]
```

```{r}
res.Batch.GeneSel.pair <- mod.t.test(x = normDataBatch.GeneSel, 
                                     group = SampleInfo$group,
                                     paired = TRUE, 
                                     subject = SampleInfo$cohort)
#res.Batch.GeneSel.pair
colSums(res.Batch.GeneSel.pair[,c(5,6)] < 0.05)
# results: 29 x p.val & 0 x adj.p.val 

sig.p <- row.names(res.Batch.GeneSel.pair)[which(res.Batch.GeneSel.pair[,5]<0.05)]
sig.p
GeneID.sig.p <- GeneID.GeneSel[GeneID.GeneSel$PROBEID %in% sig.p,]
GeneID.sig.p
unique(GeneID.sig.p$SYMBOL)
```


### Export of Results, Gene list

```{r}
res.GeneSel.GSE36398 <- as.data.frame(res.Batch.GeneSel.pair)
res.GeneSel.GSE36398$PROBEID <- ProbeID.GeneSel
res.GeneSel.GSE36398 <- merge(res.GeneSel.GSE36398, GeneID.GeneSel, by = "PROBEID", all.x = TRUE)
res.GeneSel.GSE36398<- res.GeneSel.GSE36398[order(res.GeneSel.GSE36398$p.value),]
write.csv(res.GeneSel.GSE36398, file = "ResultsGeneSel_GSE36398.csv", row.names = FALSE)
save(res.GeneSel.GSE36398, file = "ResultsGeneSel_GSE36398.RData")
```


## All genes

### mod. t-test

```{r}
res <- mod.t.test(x = normDataBatch, group = SampleInfo$group)
colSums(res[,c(5,6)] < 0.05)
# no results for adj. p value < 5%
```


# Used software

```{r}
sessionInfo()
```

