---
title: "Analysis of GSE56787 celllines"
author: "T. Schaetzl and M. Kohl"
date: "August 2022"
output:
  rmarkdown::html_document:
    theme: united
    highlight: tango
    toc: true
---

# Loading of required packages

```{r}
library(tximport)
library(MKomics)
library(vsn)
library(limma)
library(MKinfer)
library(VennDiagram)
library(edgeR)
library(GO.db)
library(org.Hs.eg.db)
```


# Import of Salmon Quantification data

```{r}
SampleInfo <- read.csv("GSE56787_SampleInfo.csv", na.strings = "N/A")
SampleInfo
txi <- readRDS(file = "SRP043221.rds")
head(txi$counts)
cts <- txi$counts
cts <- cts[,order(colnames(cts))]
cts <- cts[,SampleInfo$Experiment_type=='cell_lines']
SampleInfo <- SampleInfo[SampleInfo$Experiment_type=='cell_lines',]
sort(colSums(cts)/1e6)
```


# Quality control of raw data

```{r}
cts0 <- cts
cts0[cts0 == 0] <- min(cts0[cts0 != 0])
boxplot(cts0, log = "y", range = 0, las = 2)
```

```{r}
Cor <- cor(cts, method = "spearman")
corPlot2(Cor, minCor = min(Cor), 
         labels = paste(SampleInfo$SRR),
         title = "Spearman correlation")
```


# Preprocessing

```{r}
dge <- DGEList(counts=cts)
group <- SampleInfo$group
celltype <- factor(SampleInfo$cell_type)
design <- model.matrix(~ group*celltype)
keep <- filterByExpr(dge, design)
dge <- dge[keep,,keep.lib.sizes=FALSE]
dge <- calcNormFactors(dge)
logCPM <- cpm(dge, log=TRUE, prior.count=3)
```


## Quality control of filtered data

```{r}
boxplot(logCPM, range = 0, las = 2)
```

```{r}
Cor <- cor(logCPM, method = "spearman")
corPlot2(Cor, minCor = min(Cor), 
         labels=paste(SampleInfo$SRR),
         title = "Spearman correlation")
```

```{r}
meanSdPlot(logCPM)
```

```{r}
library(org.Hs.eg.db)
symbols <- sapply(strsplit(rownames(logCPM), "\\|"), "[", 6)
ensemblIDs <- sapply(strsplit(rownames(logCPM), "\\|"), "[", 2)
ensemblIDs <- sapply(strsplit(ensemblIDs, "\\."), "[", 1)
entrezIDs <- mapIds(org.Hs.eg.db, keys = ensemblIDs, column = "ENTREZID", 
                    keytype = "ENSEMBL", multiVals = "first")
```

```{r}
ExportData <- t(logCPM)
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE56787data_celllines.rds")
```


# Statistical Analysis

## Gene list

```{r}
library(openxlsx)
GeneSel <- read.xlsx("FSHDlist.xlsx")
GeneSel$ENSEMBL <- sapply(strsplit(GeneSel$gene_id, "\\."), "[", 1)
dim(GeneSel)
```

```{r}
logCPM.Genesel <- logCPM[ensemblIDs %in% GeneSel$ENSEMBL,]
GeneSel[!(GeneSel$ENSEMBL %in% ensemblIDs),]
```

```{r}
library(org.Hs.eg.db)
symbols.Genesel <- sapply(strsplit(rownames(logCPM.Genesel), "\\|"), "[", 6)
ensemblIDs.Genesel <- sapply(strsplit(rownames(logCPM.Genesel), "\\|"), "[", 2)
ensemblIDs.Genesel <- sapply(strsplit(ensemblIDs.Genesel, "\\."), "[", 1)
entrezIDs.Genesel <- mapIds(org.Hs.eg.db, keys = ensemblIDs.Genesel, column = "ENTREZID", 
                    keytype = "ENSEMBL", multiVals = "first")
```

```{r}
library(MKomics)
fit <- lmFit(logCPM.Genesel,design)
fit <- eBayes(fit)

## group
resgroup <- topTable(fit, coef="groupFSHD", number = Inf)
sum(resgroup$P.Value < 0.05)
# 37 results
sort(unique(symbols.Genesel[resgroup$P.Val < 0.05]))
sum(resgroup$adj.P.Value < 0.05)
# no results

## celltype
rescelltype <- topTable(fit, coef="celltypeMT", number = Inf)
sum(rescelltype$P.Value < 0.05)
# 26 results
sort(unique(symbols.Genesel[rescelltype$P.Val < 0.05]))
sum(rescelltype$adj.P.Val < 0.05)
# 1 result
sort(unique(symbols.Genesel[rescelltype$adj.P.Val < 0.05]))

## group:celltype
rescelltypegroup <- topTable(fit, coef="groupFSHD:celltypeMT", number = Inf)
sum(rescelltypegroup$P.Value < 0.05)
# 74 results
sort(unique(symbols.Genesel[rescelltypegroup$P.Val < 0.05]))
sum(rescelltypegroup$adj.P.Val < 0.05)
# 19 results
sort(unique(symbols.Genesel[rescelltypegroup$adj.P.Val < 0.05]))
```


## Export of Results, Gene list

```{r}
temp1 <- resgroup
colnames(temp1) <- paste0("group: ", colnames(resgroup))
temp2 <- rescelltype
colnames(temp2) <- paste0("celltype: ", colnames(rescelltype))
temp3 <- rescelltypegroup
colnames(temp3) <- paste0("group:celltype: ", 
                                      colnames(rescelltypegroup))
temp1.ID <- sapply(strsplit(rownames(temp1), "\\|"), "[", 2)
temp1.ID <- sapply(strsplit(temp1.ID, "\\."), "[", 1)
temp2.ID <- sapply(strsplit(rownames(temp2), "\\|"), "[", 2)
temp2.ID <- sapply(strsplit(temp2.ID, "\\."), "[", 1)
temp3.ID <- sapply(strsplit(rownames(temp3), "\\|"), "[", 2)
temp3.ID <- sapply(strsplit(temp3.ID, "\\."), "[", 1)
res.GeneSel.GSE56787celllines <- cbind(temp1[order(temp1.ID),], temp2[order(temp2.ID),], temp3[order(temp3.ID),])
res.GeneSel.GSE56787celllines$ENSEMBLE <- sort(temp1.ID)
rm(temp1,temp2, temp3)
write.csv(res.GeneSel.GSE56787celllines, file = "ResultsGeneSel_GSE56787celllines.csv")
save(res.GeneSel.GSE56787celllines, file = "ResultsGeneSel_GSE56787celllines.RData")
```

## All genes

```{r}
library(MKomics)
design <- model.matrix(~ group * celltype)
fit <- lmFit(logCPM,design)
fit <- eBayes(fit)

## group
resgroup <- topTable(fit, coef="groupFSHD", number = Inf)
sum(resgroup$P.Value < 0.05)
sum(resgroup$adj.P.Value < 0.05)
# no results for adj. p value < 5%

## celltype
rescelltype <- topTable(fit, coef="celltypeMT", number = Inf)
sum(rescelltype$P.Value < 0.05)
sum(rescelltype$adj.P.Val < 0.05)
# 295 results for adj. p value < 5%
sort(unique(symbols[rescelltype$adj.P.Val < 0.05]))

## group:celltype
rescelltypegroup <- topTable(fit, coef="groupFSHD:celltypeMT", number = Inf)
sum(rescelltypegroup$P.Value < 0.05)
sum(rescelltypegroup$adj.P.Val < 0.05) 
# 67 results for adj. p value < 5%
sort(unique(symbols[rescelltypegroup$adj.P.Val < 0.05]))
```


## Export of Results

```{r}
temp1 <- resgroup
colnames(temp1) <- paste0("group: ", colnames(resgroup))
temp2 <- rescelltype
colnames(temp2) <- paste0("celltype: ", colnames(rescelltype))
temp3 <- rescelltypegroup
colnames(temp3) <- paste0("group:celltype: ", 
                                      colnames(rescelltypegroup))
res.GSE56787celllines <- cbind(temp1, temp2, temp3, ensemblIDs)
rm(temp1,temp2, temp3)
names(res.GSE56787celllines)[19] <- "ENSEMBLE"
write.csv(res.GSE56787celllines, file = "Results_GSE56787celllines.csv")
save(res.GSE56787celllines, file = "Results_GSE56787celllines.RData")
```

## Heatmap 

```{r}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
## Data selection
sel <- rescelltypegroup$adj.P.Val < 0.05
selData <- logCPM[sel,]
selData <- selData - rowMeans(selData, na.rm = TRUE)
colnames(selData) <- paste(SampleInfo$subject)
## Color definition
col1 <- rev(brewer.pal(n = 8, name = "RdYlBu"))
col2 <- brewer.pal(n = 3, name = "Set1")[c(3,1)]
col3 <- brewer.pal(n = 3, name = "Set2")[c(3,1)]
col4 <- brewer.pal(n = 3, name = "Dark2")[c(2,4)]
mycol <- colorRamp2(seq(-2, 2, length = 128), 
                    colorRampPalette(col1)(128))
## Annotation
df <- data.frame(group = SampleInfo$group)
ha <- HeatmapAnnotation(df = df,
                        col = list(group = c("control" = col2[1],
                                              "FSHD" = col2[2])))
## Drawing the heatmap
Heatmap(selData, col = mycol, name = "log2-FC", show_row_names = FALSE,
        use_raster = FALSE,
        top_annotation = ha, show_column_names = TRUE,
        column_names_gp = gpar(fontsize = 8),
        clustering_distance_columns = "euclidean",
        clustering_distance_rows = "euclidean",
        show_row_dend = TRUE,
        cluster_columns = TRUE, show_column_dend = TRUE,
        column_title = "p value < 0.05 and |log2-FC| > log2(1.5)",
        show_heatmap_legend = TRUE)
```


### Enrichment Analysis for group:celltype

```{r}

GeneID.UP <- entrezIDs[(rescelltypegroup$logFC > 0 & 
                        rescelltypegroup$adj.P.Val < 0.05)]
GeneID.DN <- entrezIDs[(rescelltypegroup$logFC < 0 & 
                        rescelltypegroup$adj.P.Val < 0.05)]


symbols[(rescelltypegroup$logFC > 0 & 
                        rescelltypegroup$adj.P.Val < 0.05)]
```


#### Calculation of enrichment for GO

```{r}
library(GO.db)
go <- goana(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:20,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:20,]
```


### Export of Results: GO

```{r}
go.up.GSE56787celllines <- go.up
go.up.GSE56787celllines <- go.up.GSE56787celllines[order(go.up.GSE56787celllines$P.Up),]
write.csv(go.up.GSE56787celllines, file = "GO.UP_GSE56787celllines_rescelltypegroup.csv", row.names = FALSE)
save(go.up.GSE56787celllines, file = "GO.UP_GSE56787celllines_rescelltypegroup.RData")
```

```{r}
go.dn.GSE56787celllines <- go.dn
go.dn.GSE56787celllines <- go.dn.GSE56787celllines[order(go.dn.GSE56787celllines$P.Down),]
write.csv(go.dn.GSE56787celllines, file = "GO.DN_GSE56787celllines_rescelltypegroup.csv", row.names = FALSE)
save(go.dn.GSE56787celllines, file = "GO.DN_GSE56787celllines_rescelltypegroup.RData")
```


#### Calculation of enrichment for KEGG

```{r}
kg <- kegga(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
kegg.up <- topKEGG(kg, sort = "up", number = Inf)
kegg.up[1:20,]
kegg.dn <- topKEGG(kg, sort = "down", number = Inf)
kegg.dn[1:20,]
```


### Export of Results: KEGG

```{r}
kegg.up.GSE56787 <- kegg.up
kegg.up.GSE56787 <- kegg.up.GSE56787[order(kegg.up.GSE56787$P.Up),]
write.csv(kegg.up.GSE56787, file = "KEGG.UP_GSE56787celllines_rescelltypegroup.csv", row.names = FALSE)
save(kegg.up.GSE56787, file = "KEGG.UP_GSE56787celllines_rescelltypegroup.RData")
```

```{r}
kegg.dn.GSE56787 <- kegg.dn
kegg.dn.GSE56787 <- kegg.dn.GSE56787[order(kegg.dn.GSE56787$P.Down),]
write.csv(kegg.dn.GSE56787, file = "KEGG.DN_GSE56787celllines_rescelltypegroup.csv", row.names = FALSE)
save(kegg.dn.GSE56787, file = "KEGG.DN_GSE56787celllines_rescelltypegroup.RData")
```


# Used software

```{r}
sessionInfo()
```

