---
title: "Random-Effects Meta-Analysis"
author: "T. Schätzl and M. Kohl"
date: "December 2022"
output:
  rmarkdown::html_document:
    toc: true
    theme: united
    highlight: tango
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Load summarized data

```{r}
## read all RData files from folder summary
files.all <- list.files(path = "../summary", full.names = TRUE)
files.RData <- files.all[grep("RData", files.all)]
for(i in 1:length(files.RData)){
  load(files.RData[i])
}
```


## Merge summarized data

```{r}
###############################################################################
## GSE10760
###############################################################################
tmp <- res.GSE10760[,c("ENSEMBL", "mean of control", "mean of FSHD",
                       "SD of control", "SD of FSHD", "n of control", 
                       "n of FSHD")]
names(tmp)[2:7] <- c("GSE10760.Mctrl", "GSE10760.Mfshd", 
                     "GSE10760.SDctrl", "GSE10760.SDfshd",
                     "GSE10760.nctrl", "GSE10760.nfshd")
resData <- tmp


###############################################################################
## GSE15090
###############################################################################
tmp <- res.GSE15090[,c("ENSEMBL", "mean of control", "mean of FSHD",
                       "SD of control", "SD of FSHD", "n of control",
                       "n of FSHD")]
names(tmp)[2:7] <- c("GSE15090.Mctrl", "GSE15090.Mfshd", 
                     "GSE15090.SDctrl", "GSE15090.SDfshd",
                     "GSE15090.nctrl", "GSE15090.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## GSE26061
###############################################################################
tmp <- res.GSE26061[,c("ENSEMBL", "mean of control", "mean of FSHD",
                       "SD of control", "SD of FSHD", "n of control",
                       "n of FSHD")]
names(tmp)[2:7] <- c("GSE26061.Mctrl", "GSE26061.Mfshd", 
                     "GSE26061.SDctrl", "GSE26061.SDfshd",
                     "GSE26061.nctrl", "GSE26061.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## GSE26145
###############################################################################
tmp <- res.GSE26145[,c("ENSEMBL", "mean of control", "mean of FSHD",
                       "SD of control", "SD of FSHD", "n of control",
                       "n of FSHD")]
names(tmp)[2:7] <- c("GSE26145.Mctrl", "GSE26145.Mfshd", 
                     "GSE26145.SDctrl", "GSE26145.SDfshd",
                     "GSE26145.nctrl", "GSE26145.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## GSE26852
###############################################################################
tmp <- res.GSE26852[,c("ENSEMBL", "mean of control", "mean of FSHD",
                       "SD of control", "SD of FSHD", "n of control",
                       "n of FSHD")]
names(tmp)[2:7] <- c("GSE26852.Mctrl", "GSE26852.Mfshd", 
                     "GSE26852.SDctrl", "GSE26852.SDfshd",
                     "GSE26852.nctrl", "GSE26852.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## GSE36398
###############################################################################
tmp <- res.GSE36398[,c("ENSEMBL", "mean of control", "mean of FSHD",
                       "SD of control", "SD of FSHD", "n of control",
                       "n of FSHD")]
names(tmp)[2:7] <- c("GSE36398.Mctrl", "GSE36398.Mfshd", 
                     "GSE36398.SDctrl", "GSE36398.SDfshd",
                     "GSE36398.nctrl", "GSE36398.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## GSE56787
###############################################################################
tmp <- res.GSE56787[,c("ENSEMBL", "mean of control", "mean of FSHD",
                       "SD of control", "SD of FSHD", "n of control",
                       "n of FSHD")]
names(tmp)[2:7] <- c("GSE56787.Mctrl", "GSE56787.Mfshd", 
                     "GSE56787.SDctrl", "GSE56787.SDfshd",
                     "GSE56787.nctrl", "GSE56787.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## GSE56787 cellines
###############################################################################
tmp <- res.GSE56787celllines[,c("ENSEMBL", "mean of control", "mean of FSHD",
                       "SD of control", "SD of FSHD", "n of control",
                       "n of FSHD")]
names(tmp)[2:7] <- c("GSE56787celllines.Mctrl", "GSE56787celllines.Mfshd", 
                     "GSE56787celllines.SDctrl", "GSE56787celllines.SDfshd",
                     "GSE56787celllines.nctrl", "GSE56787celllines.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## GSE115650
###############################################################################
tmp <- res.GSE115650[,c("ENSEMBL", "mean of control", "mean of FSHD",
                       "SD of control", "SD of FSHD", "n of control",
                       "n of FSHD")]
names(tmp)[2:7] <- c("GSE115650.Mctrl", "GSE115650.Mfshd", 
                     "GSE115650.SDctrl", "GSE115650.SDfshd",
                     "GSE115650.nctrl", "GSE115650.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## GSE123468_12
###############################################################################
tmp <- res.GSE123468_12[,c("ENSEMBL", "mean of control", "mean of FSHD",
                        "SD of control", "SD of FSHD", "n of control",
                        "n of FSHD")]
names(tmp)[2:7] <- c("GSE123468_12.Mctrl", "GSE123468_12.Mfshd", 
                     "GSE123468_12.SDctrl", "GSE123468_12.SDfshd",
                     "GSE123468_12.nctrl", "GSE123468_12.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## GSE123468_16
###############################################################################
tmp <- res.GSE123468_16[,c("ENSEMBL", "mean of control", "mean of FSHD",
                           "SD of control", "SD of FSHD", "n of control",
                           "n of FSHD")]
names(tmp)[2:7] <- c("GSE123468_16.Mctrl", "GSE123468_16.Mfshd", 
                     "GSE123468_16.SDctrl", "GSE123468_16.SDfshd",
                     "GSE123468_16.nctrl", "GSE123468_16.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## EGAD00001008337
###############################################################################
tmp <- res.EGA[,c("ENSEMBL", "mean of control", "mean of FSHD",
                  "SD of control", "SD of FSHD", "n of control",
                  "n of FSHD")]
names(tmp)[2:7] <- c("EGAD00001008337.Mctrl", "EGAD00001008337.Mfshd", 
                     "EGAD00001008337.SDctrl", "EGAD00001008337.SDfshd",
                     "EGAD00001008337.nctrl", "EGAD00001008337.nfshd")
resData <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)


###############################################################################
## GSE138768 (DUX4)
###############################################################################
tmp <- res.GSE138768[,c("ENSEMBL", "mean of control", "mean of DUX4",
                        "SD of control", "SD of DUX4", "n of control",
                        "n of DUX4")]
names(tmp)[2:7] <- c("GSE138768.Mctrl", "GSE138768.Mfshd",
                     "GSE138768.SDctrl", "GSE138768.SDfshd",
                     "GSE138768.nctrl", "GSE138768.nfshd")
resData.D4 <- merge(resData, tmp, by = "ENSEMBL", all = TRUE)
```


## Get annotation information

```{r}
library(AnnotationHub)
ah <- AnnotationHub()
query(ah, "EnsDb")
ahDb <- query(ah, pattern = c("Homo sapiens", "EnsDb", 108))
ahEdb <- ahDb[[1]]
gns <- genes(ahEdb)
gnsIDs <- gns@elementMetadata@listData
resData$gene_name <- ""
resData$entrezid <- ""
for(i in 1:nrow(resData)){
  ensembleIDs <- unlist(strsplit(resData$ENSEMBL[i], " \\| "))
  sel <- which(gnsIDs$gene_id %in% ensembleIDs)
  gene.name <- gnsIDs$gene_name[sel]
  if(length(gene.name) > 0) gene.name <- gene.name[gene.name != ""]
  resData$gene_name[i] <- paste(sort(unique(gene.name)), collapse = " | ")
  entrezID <- gnsIDs$entrezid[sel]
  entrezID <- unlist(entrezID[!is.na(entrezID)])
  resData$entrezid[i] <- paste(sort(unique(entrezID)), collapse = " | ")
}
```


## Filter

```{r}
## only genes that were measured at least three times
resData <- resData[which(rowSums(!is.na(resData[,2:73])) >= 18),]
```

```{r}
library(metafor)
effs <- numeric(nrow(resData))

ind.Mctrl <- grep("Mctrl", names(resData))
ind.Mfshd <- grep("Mfshd", names(resData))
ind.SDctrl <- grep("SDctrl", names(resData))
ind.SDfshd <- grep("SDfshd", names(resData))
ind.nctrl <- grep("nctrl", names(resData))
ind.nfshd <- grep("nfshd", names(resData))
for(i in 1:nrow(resData)){
  tmp <- escalc(measure = "SMDH", 
                m1i = unlist(resData[i,ind.Mfshd]),
                m2i = unlist(resData[i,ind.Mctrl]),
                sd1i = unlist(resData[i,ind.SDfshd]),
                sd2i = unlist(resData[i,ind.SDctrl]),
                n1i = unlist(resData[i,ind.nfshd]),
                n2i = unlist(resData[i,ind.nctrl]))
  fit <- try(rma(yi, vi, data = tmp, measure = "SMDH", method = "REML"))
  if(inherits(fit, "try-error")){
    effs[i] <- NA
  }else{
    effs[i] <- fit$beta[1,1]
  }
}
## at least small effect size
resData <- resData[which(abs(effs) > 0.2),]
```


## Meta-analysis

```{r}
library(metafor)
library(metasens)
pvals <- numeric(nrow(resData))
effs <- numeric(nrow(resData))
CIlow <- numeric(nrow(resData))
CIupp <- numeric(nrow(resData))
I2 <- numeric(nrow(resData))
pvals.fix <- numeric(nrow(resData))
k <- numeric(nrow(resData))
egger.pval <- numeric(nrow(resData))
trimfill.side <- character(nrow(resData))
trimfill.nr <- numeric(nrow(resData))
lfk.res <- character(nrow(resData))


ind.Mctrl <- grep("Mctrl", names(resData))
ind.Mfshd <- grep("Mfshd", names(resData))
ind.SDctrl <- grep("SDctrl", names(resData))
ind.SDfshd <- grep("SDfshd", names(resData))
ind.nctrl <- grep("nctrl", names(resData))
ind.nfshd <- grep("nfshd", names(resData))
for(i in 1:nrow(resData)){
  tmp <- escalc(measure = "SMDH", 
                m1i = unlist(resData[i,ind.Mfshd]),
                m2i = unlist(resData[i,ind.Mctrl]),
                sd1i = unlist(resData[i,ind.SDfshd]),
                sd2i = unlist(resData[i,ind.SDctrl]),
                n1i = unlist(resData[i,ind.nfshd]),
                n2i = unlist(resData[i,ind.nctrl]))
  ## random-effects meta-analysis
  fit <- try(rma(yi, vi, data = tmp, measure = "SMDH", method = "REML"))
  if(inherits(fit, "try-error")){
    pvals[i] <- NA
    effs[i] <- NA
    CIlow[i] <- NA
    CIupp[i] <- NA
    I2[i] <- NA
    k[i] <- NA
    egger.pval[i] <- NA
    trimfill.side[i] <- NA
    trimfill.nr[i] <- NA
    lfk.res[i] <- NA
  }else{
    pvals[i] <- fit$pval
    effs[i] <- fit$beta[1,1]
    CIlow[i] <- fit$ci.lb
    CIupp[i] <- fit$ci.ub
    I2[i] <- fit$I2
    k[i] <- fit$k
    ## Egger test
    res.regtest <- try(regtest(fit, model = "lm"))
    if(inherits(res.regtest, "try-error")){
      egger.pval[i] <- NA
    }else{
      egger.pval[i] <- res.regtest$pval
    }
    ## trim-and-fill analysis
    tf.fit <- try(trimfill(fit))
    if(inherits(tf.fit, "try-error")){
      trimfill.side[i] <- NA
      trimfill.nr[i] <- NA
    }else{
      trimfill.side[i] <- tf.fit$side
      trimfill.nr[i] <- tf.fit$k0
    }
    ## LFK index
    lfk <- lfkindex(yi, sqrt(vi), data = tmp)
    lfk.res[i] <- lfk$interpretation
  }
  ## fixed-effects meta-analysis
  fit2 <- try(rma(yi, vi, data = tmp, measure = "SMDH", method = "FE"))
  if(inherits(fit2, "try-error")){
    pvals.fix[i] <- NA
  }else{
    pvals.fix[i] <- fit2$pval
  }
}

## add results to resData
resData$p.value <- pvals
resData$eff <- effs
resData$CI.low <- CIlow
resData$CI.upp <- CIupp
resData$I2 <- I2
resData$p.value.fix <- pvals.fix
resData$k <- k
resData$egger.pval <- egger.pval
resData$trimfill.side <- trimfill.side
resData$trimfill.nr <- trimfill.nr
resData$lfk.res <- lfk.res
## adjust p values by FDR
resData$adj.p.value <- p.adjust(resData$p.value, method = "fdr")
resData$adj.p.value.fix <- p.adjust(resData$p.value.fix, method = "fdr")
## add DUX4 dataset
resData <- merge(resData, resData.D4[,c(1,74:79)], by = "ENSEMBL", all.x = TRUE)
```


## Export results

```{r}
save(resData, file = "Results_meta.RData")
```


## Investigation of heterogeneity

```{r}
## might not be important
sum(resData$I2 <= 40)
## moderate heterogeneity
sum(resData$I2 >= 30 & resData$I2 <= 60)
## substantial heterogeneity
sum(resData$I2 >= 50 & resData$I2 <= 90)
## considerable heterogeneity
sum(resData$I2 >= 75)
```


## Investigation of publication bias

### Regression test by Egger

```{r}
## raw p-values
sum(resData$egger.pval < 0.05)
## FDR adjusted p-values
sum(p.adjust(resData$egger.pval, method = "fdr") < 0.05)
```


### Trim-and-fill analysis

```{r}
library(DescTools)
## all meta-analyses
Freq(as.factor(resData$trimfill.nr), useNA = "ifany")
## significant meta-analyses
Freq(as.factor(resData$trimfill.nr[which(resData$adj.p.value < 0.05)]), useNA = "ifany")
```


### LFK index

```{r}
## all meta-analyses
Freq(resData$lfk.res, useNA = "ifany")
## significant meta-analyses
Freq(resData$lfk.res[which(resData$adj.p.value < 0.05)], useNA = "ifany")
```


## Number of significant results

### Random-effects meta-analysis

```{r}
## raw p-values
sum(resData$p.value < 0.05)
## FDR adjusted p-values
sum(resData$adj.p.value < 0.05)

```


### Fixed-effects meta-analysis

```{r}
## raw p-values
sum(resData$p.value.fix < 0.05)
## FDR adjusted p-values
sum(resData$adj.p.value.fix < 0.05)
```


## Draw heatmap

```{r}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
## Preparation of data
resData.sig <- resData[which(resData$adj.p.value < 0.05),]
DS <- unique(sapply(strsplit(names(resData.sig)[c(2:73,89:94)], "\\."), "[", 1))
diffData <- matrix(ncol = length(DS), nrow = nrow(resData.sig))
colnames(diffData) <- DS
ind.Mctrl <- grep("Mctrl", names(resData.sig))
ind.Mfshd <- grep("Mfshd", names(resData.sig))
ind.SDctrl <- grep("SDctrl", names(resData.sig))
ind.SDfshd <- grep("SDfshd", names(resData.sig))
ind.nctrl <- grep("nctrl", names(resData.sig))
ind.nfshd <- grep("nfshd", names(resData.sig))
for(i in 1:nrow(diffData)){
  tmp <- escalc(measure = "SMDH", 
                m1i = unlist(resData.sig[i,ind.Mfshd]),
                m2i = unlist(resData.sig[i,ind.Mctrl]),
                sd1i = unlist(resData.sig[i,ind.SDfshd]),
                sd2i = unlist(resData.sig[i,ind.SDctrl]),
                n1i = unlist(resData.sig[i,ind.nfshd]),
                n2i = unlist(resData.sig[i,ind.nctrl]))
  diffData[i,] <- tmp$yi
}
diffData <- cbind(diffData, resData.sig$eff)
colnames(diffData)[14] <- "Meta-Analysis"
rownames(diffData) <- resData.sig$gene_name
selData <- diffData[order(diffData[,14], decreasing = TRUE),]
selData <- selData[,c(13,1:12,14)]
colnames(selData)[1] <- paste(colnames(selData)[1], "(DUX4)")
## prepartion of heatmap
col1 <- rev(brewer.pal(n = 11, name = "RdYlBu"))
mycol <- colorRamp2(seq(-2, 2, length = 128), colorRampPalette(col1)(128))
col2 <- hcl.colors(10)
df <- data.frame("Technology" = factor(c("HiSeq 2500", "HG-U133A+B", "HG-U133_Plus_2", "HuEx10stv2_Hs_ENSG",
                                "HuEx-1_0-st", "HumanHT-12 V3.0", "HuGene-1_0-st",
                                "HiSeq 2500", "HiSeq 2500", "HiSeq 2500",
                                "HiSeq 2500", "HiSeq 2500", "NextSeq 500", "Meta-Analysis"), 
                                levels = c("HG-U133A+B", "HG-U133_Plus_2", "HuEx10stv2_Hs_ENSG",
                                           "HuEx-1_0-st", "HumanHT-12 V3.0", "HuGene-1_0-st",
                                           "HiSeq 2500", "NextSeq 500", "Meta-Analysis")))
ha <- HeatmapAnnotation(df = df, 
                        col = list("Technology" = factor(c("HG-U133A+B" = col2[1], 
                                                  "HG-U133_Plus_2" = col2[2], 
                                                  "HuEx10stv2_Hs_ENSG" = col2[3],
                                                  "HuEx-1_0-st" = col2[4], 
                                                  "HumanHT-12 V3.0" = col2[5], 
                                                  "HuGene-1_0-st" = col2[6],
                                                  "HiSeq 2500" = col2[7], 
                                                  "NextSeq 500" = col2[8],
                                                  "Meta-Analysis" = col2[9]))),
                        show_annotation_name = FALSE,
                        annotation_legend_param = list(nrow = 2, title = "", 
                                                       labels_gp = gpar(fontsize = 12)))
## generate heatmap
hm <- Heatmap(selData, col = mycol, top_annotation = ha, na_col = "grey",
        heatmap_legend_param = list(title = "STD\nlog2-FC",
                                    title_gp = gpar(fontsize = 12, fontface = "bold"),
                                    legend_height = unit(7, "cm"),
                                    at = seq(-2, 2, by = 0.5)),
        column_title = rep("", 3),
        show_row_names = FALSE,
        row_names_gp = gpar(fontsize = 3), use_raster = FALSE,
        cluster_rows = FALSE, cluster_columns = FALSE,
        column_split = factor(c("DUX4", rep("Technology", 12), "Meta-Analysis"),
                              levels = c("DUX4", "Technology", "Meta-Analysis")),
        row_split = factor(c(rep("FSHD up", sum(selData[,14] > 0)),
                      rep("FSHD down", sum(selData[,14] < 0))),
                      levels = c("FSHD up", "FSHD down")))
draw(hm, annotation_legend_side = "bottom")
```


## Export data of heatmap

```{r}
library(openxlsx)
hmData <- data.frame(diffData, p_value = resData.sig$p.value,
                     adj_p_value = resData.sig$adj.p.value,
                     ENSEMBL = resData.sig$ENSEMBL, 
                     gene_name = resData.sig$gene_name, 
                     ENTREZID = resData.sig$entrezid)
hmData <- data.frame(hmData[order(hmData[,14], decreasing = TRUE),])
write.xlsx(hmData, file = "Heatmap_data.xlsx")
```


## LOO sensitivity analysis

### Dataset level

```{r}
dataSets <- unique(sapply(strsplit(names(resData), "\\."), "[", 1))[2:13]
resData.LOO <- vector(mode = "list", length = length(dataSets))
names(resData.LOO) <- dataSets
j <- 0
for(Name in dataSets){
  j <- j + 1
  cat("======================================\n")
  cat("j:\t", j, "\n")
  cat("======================================\n")
  ind.rm <- grep(Name, names(resData))
  resData.red <- resData[,-ind.rm]
  ## only genes that were measured at least three times
  resData.red <- resData.red[rowSums(!is.na(resData.red[,2:67])) >= 18,]
  ## only genes with at least a small effect (d >= 0.2)
  resData.red <- resData.red[which(abs(resData.red$eff) > 0.2),]
  pvals <- numeric(nrow(resData.red))
  effs <- numeric(nrow(resData.red))
  CIlow <- numeric(nrow(resData.red))
  CIupp <- numeric(nrow(resData.red))
  I2 <- numeric(nrow(resData.red))
  
  ind.Mctrl <- grep("Mctrl", names(resData.red))
  ind.Mfshd <- grep("Mfshd", names(resData.red))
  ind.SDctrl <- grep("SDctrl", names(resData.red))
  ind.SDfshd <- grep("SDfshd", names(resData.red))
  ind.nctrl <- grep("nctrl", names(resData.red))
  ind.nfshd <- grep("nfshd", names(resData.red))
  for(i in 1:nrow(resData.red)){
    tmp <- escalc(measure = "SMDH", 
                  m1i = unlist(resData.red[i,ind.Mfshd]),
                  m2i = unlist(resData.red[i,ind.Mctrl]),
                  sd1i = unlist(resData.red[i,ind.SDfshd]),
                  sd2i = unlist(resData.red[i,ind.SDctrl]),
                  n1i = unlist(resData.red[i,ind.nfshd]),
                  n2i = unlist(resData.red[i,ind.nctrl]))
    fit <- try(rma(yi, vi, data = tmp, measure = "SMDH", method = "REML"))
    if(inherits(fit, "try-error")){
      pvals[i] <- NA
      effs[i] <- NA
      CIlow[i] <- NA
      CIupp[i] <- NA
      I2[i] <- NA
    }else{
      pvals[i] <- fit$pval
      effs[i] <- fit$beta[1,1]
      CIlow[i] <- fit$ci.lb
      CIupp[i] <- fit$ci.ub
      I2[i] <- fit$I2
    }
  }
  resData.red$p.value <- pvals
  resData.red$adj.p.value <- p.adjust(pvals, method = "fdr")
  resData.red$eff <- effs
  resData.red$CI.low <- CIlow
  resData.red$CI.upp <- CIupp
  resData.red$I2 <- I2
  resData.LOO[[j]] <- resData.red
}

## Nunber of studies for significant genes
table(resData.sig$k)
## at least four datasets
nrow(resData.sig[resData.sig$k > 3,])

## number of significant results 
sapply(resData.LOO, function(x) sum(x$p.value < 0.05, na.rm = TRUE))
sapply(resData.LOO, function(x) sum(x$adj.p.value < 0.05, na.rm = TRUE))

## overlap between LOO results
resData.LOO.sig <- lapply(resData.LOO, function(x) x[x$p.value < 0.05,c("ENSEMBL", "p.value", "adj.p.value")])
intersect.sig <- resData.LOO.sig[[1]]$ENSEMBL
for(i in 2:length(resData.LOO.sig)){
  intersect.sig <- intersect(intersect.sig, resData.LOO.sig[[i]]$ENSEMBL)
}
length(intersect(intersect.sig, resData.sig$ENSEMBL))
length(intersect(intersect.sig, resData.sig$ENSEMBL))/nrow(resData.sig[resData.sig$k > 3,])
```


### Study level

```{r}
studies <- unique(sapply(strsplit(names(resData), "\\."), "[", 1))[2:13]
studies[c(7,10)] <- c("GSE56787", "GSE123468")
studies <- studies[-c(8,11)]
resData.LOO2 <- vector(mode = "list", length = length(studies))
names(resData.LOO2) <- studies
j <- 0
for(Name in studies){
  j <- j + 1
  cat("======================================\n")
  cat("j:\t", j, "\n")
  cat("======================================\n")
  ind.rm <- grep(Name, names(resData))
  resData.red <- resData[,-ind.rm]
  ## only genes that were measured at least three times
  resData.red <- resData.red[rowSums(!is.na(resData.red[,2:67])) >= 18,]
  ## only genes with at least a small effect (d >= 0.2)
  resData.red <- resData.red[which(abs(resData.red$eff) > 0.2),]
  pvals <- numeric(nrow(resData.red))
  effs <- numeric(nrow(resData.red))
  CIlow <- numeric(nrow(resData.red))
  CIupp <- numeric(nrow(resData.red))
  I2 <- numeric(nrow(resData.red))
  
  ind.Mctrl <- grep("Mctrl", names(resData.red))
  ind.Mfshd <- grep("Mfshd", names(resData.red))
  ind.SDctrl <- grep("SDctrl", names(resData.red))
  ind.SDfshd <- grep("SDfshd", names(resData.red))
  ind.nctrl <- grep("nctrl", names(resData.red))
  ind.nfshd <- grep("nfshd", names(resData.red))
  for(i in 1:nrow(resData.red)){
    tmp <- escalc(measure = "SMDH", 
                  m1i = unlist(resData.red[i,ind.Mfshd]),
                  m2i = unlist(resData.red[i,ind.Mctrl]),
                  sd1i = unlist(resData.red[i,ind.SDfshd]),
                  sd2i = unlist(resData.red[i,ind.SDctrl]),
                  n1i = unlist(resData.red[i,ind.nfshd]),
                  n2i = unlist(resData.red[i,ind.nctrl]))
    fit <- try(rma(yi, vi, data = tmp, measure = "SMDH", method = "REML"))
    if(inherits(fit, "try-error")){
      pvals[i] <- NA
      effs[i] <- NA
      CIlow[i] <- NA
      CIupp[i] <- NA
      I2[i] <- NA
    }else{
      pvals[i] <- fit$pval
      effs[i] <- fit$beta[1,1]
      CIlow[i] <- fit$ci.lb
      CIupp[i] <- fit$ci.ub
      I2[i] <- fit$I2
    }
  }
  resData.red$p.value <- pvals
  resData.red$adj.p.value <- p.adjust(pvals, method = "fdr")
  resData.red$eff <- effs
  resData.red$CI.low <- CIlow
  resData.red$CI.upp <- CIupp
  resData.red$I2 <- I2
  resData.LOO2[[j]] <- resData.red
}

## number of significant results 
sapply(resData.LOO2, function(x) sum(x$p.value < 0.05, na.rm = TRUE))
sapply(resData.LOO2, function(x) sum(x$adj.p.value < 0.05, na.rm = TRUE))

## overlap between LOO results
resData.LOO2.sig <- lapply(resData.LOO2, function(x) x[x$p.value < 0.05,c("ENSEMBL", "p.value", "adj.p.value")])
intersect.sig2 <- resData.LOO2.sig[[1]]$ENSEMBL
for(i in 2:length(resData.LOO2.sig)){
  intersect.sig2 <- intersect(intersect.sig2, resData.LOO2.sig[[i]]$ENSEMBL)
}
length(intersect(intersect.sig2, resData.sig$ENSEMBL))
length(intersect(intersect.sig2, resData.sig$ENSEMBL))/nrow(resData.sig[resData.sig$k > 3,])
```


## Enrichment analysis

### Extract ENTREZ IDs

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
## extract ENTREZ IDs
entrez.list <- unique(unlist(strsplit(resData$entrezid, " \\| ")))
entrez.list <- entrez.list[order(as.numeric(entrez.list))]
entrez.list <- entrez.list[!is.na(entrez.list)]
## significant up
entrez.sig <- unique(unlist(strsplit(hmData$ENTREZID[hmData$Meta.Analysis > 0], " \\| ")))
entrez.sig <- entrez.sig[order(as.numeric(entrez.sig))]
entrez.up <- entrez.sig[!is.na(entrez.sig)]
## significant down
entrez.sig <- unique(unlist(strsplit(hmData$ENTREZID[hmData$Meta.Analysis < 0], " \\| ")))
entrez.sig <- entrez.sig[order(as.numeric(entrez.sig))]
entrez.dn <- entrez.sig[!is.na(entrez.sig)]
```


### GO classification

```{r}
## BP
ggo.BP.up <- groupGO(gene = entrez.up, OrgDb = org.Hs.eg.db, ont = "BP",
                     level = 2, readable = TRUE)
ggo.BP.up@result[ggo.BP.up@result$Count > 0,]

ggo.BP.dn <- groupGO(gene = entrez.dn, OrgDb = org.Hs.eg.db, ont = "BP",
                     level = 2, readable = TRUE)
ggo.BP.dn@result[ggo.BP.dn@result$Count > 0,]

## CC
ggo.CC.up <- groupGO(gene = entrez.up, OrgDb = org.Hs.eg.db, ont = "CC",
                     level = 3, readable = TRUE)
ggo.CC.up@result[ggo.CC.up@result$Count > 0,]

ggo.CC.dn <- groupGO(gene = entrez.dn, OrgDb = org.Hs.eg.db, ont = "CC",
                     level = 3, readable = TRUE)
ggo.CC.dn@result[ggo.CC.dn@result$Count > 0,]

## MF
ggo.MF.up <- groupGO(gene = entrez.up, OrgDb = org.Hs.eg.db, ont = "MF",
                     level = 2, readable = TRUE)
ggo.MF.up@result[ggo.MF.up@result$Count > 0,]

ggo.MF.dn <- groupGO(gene = entrez.dn, OrgDb = org.Hs.eg.db, ont = "MF",
                     level = 2, readable = TRUE)
ggo.MF.dn@result[ggo.MF.dn@result$Count > 0,]
```


### GO enrichment

```{r}
## BP
ego.BP.up <- enrichGO(gene = entrez.up, universe = entrez.list,
                      OrgDb = org.Hs.eg.db, ont = "BP",
                      pAdjustMethod = "holm", pvalueCutoff = 0.05,
                      readable = TRUE)
head(ego.BP.up@result, n = 25)

ego.BP.dn <- enrichGO(gene = entrez.dn, universe = entrez.list,
                      OrgDb = org.Hs.eg.db, ont = "BP",
                      pAdjustMethod = "holm", pvalueCutoff = 0.05,
                      readable = TRUE)
head(ego.BP.dn@result, n = 25)

## CC
ego.CC.up <- enrichGO(gene = entrez.up, universe = entrez.list,
                      OrgDb = org.Hs.eg.db, ont = "CC",
                      pAdjustMethod = "holm", pvalueCutoff = 0.05,
                      readable = TRUE)
head(ego.CC.up@result, n = 25)

ego.CC.dn <- enrichGO(gene = entrez.dn, universe = entrez.list,
                      OrgDb = org.Hs.eg.db, ont = "CC",
                      pAdjustMethod = "holm", pvalueCutoff = 0.05,
                      readable = TRUE)
head(ego.CC.dn@result, n = 25)

## MF
ego.MF.up <- enrichGO(gene = entrez.up, universe = entrez.list,
                      OrgDb = org.Hs.eg.db, ont = "MF",
                      pAdjustMethod = "holm", pvalueCutoff = 0.05,
                      readable = TRUE)
head(ego.MF.up@result, n = 25)

ego.MF.dn <- enrichGO(gene = entrez.dn, universe = entrez.list,
                      OrgDb = org.Hs.eg.db, ont = "MF",
                      pAdjustMethod = "holm", pvalueCutoff = 0.05,
                      readable = TRUE)
head(ego.MF.dn@result, n = 25)
```


### Gene Set Enrichment Analysis

```{r}
geneList <- hmData$Meta.Analysis
entrez.id <- sapply(strsplit(hmData$ENTREZID[hmData$Meta.Analysis > 0], " \\| "), "[", 1)
names(geneList) <- entrez.id
geneList <- sort(geneList, decreasing = TRUE)

## BP
gse.BP <- gseGO(geneList = geneList, OrgDb = org.Hs.eg.db, ont = "BP",
                minGSSize = 10, maxGSSize = 500, pvalueCutoff = 0.05,
                verbose = TRUE, eps = 0)
head(gse.BP@result, n = 25)

## CC
gse.CC <- gseGO(geneList = geneList, OrgDb = org.Hs.eg.db, ont = "CC",
                minGSSize = 10, maxGSSize = 500, pvalueCutoff = 0.05,
                verbose = TRUE, eps = 0)
head(gse.CC@result, n = 25)

## MF
gse.MF <- gseGO(geneList = geneList, OrgDb = org.Hs.eg.db, ont = "MF",
                minGSSize = 10, maxGSSize = 500, pvalueCutoff = 0.05,
                verbose = TRUE, eps = 0)
head(gse.MF@result, n = 25)
```


### Clustering of GO terms

```{r}
library(simplifyEnrichment)
library(GO.db)
egGO2ALLEGS <- getFromNamespace("org.Hs.egGO2ALLEGS", "org.Hs.eg.db")
AnnotationDbi::Lkeys(egGO2ALLEGS) <- entrez.list
GeneID.PathID <- AnnotationDbi::toTable(egGO2ALLEGS)[, c("gene_id", "go_id", "Ontology")]
d <- duplicated(GeneID.PathID[, c("gene_id", "go_id")])
GeneID.PathID <- GeneID.PathID[!d, ]
GO.sig.up <- GeneID.PathID[GeneID.PathID$gene_id %in% entrez.up,]
GO.sig.dn <- GeneID.PathID[GeneID.PathID$gene_id %in% entrez.dn,]
## computations take long
sim.BP.up <- GO_similarity(GO.sig.up$go_id[GO.sig.up$Ontology == "BP"], ont = "BP")
sim.CC.up <- GO_similarity(GO.sig.up$go_id[GO.sig.up$Ontology == "CC"], ont = "CC")
sim.MF.up <- GO_similarity(GO.sig.up$go_id[GO.sig.up$Ontology == "MF"], ont = "MF")
sim.BP.dn <- GO_similarity(GO.sig.dn$go_id[GO.sig.dn$Ontology == "BP"], ont = "BP")
sim.CC.dn <- GO_similarity(GO.sig.dn$go_id[GO.sig.dn$Ontology == "CC"], ont = "CC")
sim.MF.dn <- GO_similarity(GO.sig.dn$go_id[GO.sig.dn$Ontology == "MF"], ont = "MF")
## computations take long
res.CC.up <- simplifyGO(sim.CC.up, column_title = "GO CC - FSHD up")
res.MF.up <- simplifyGO(sim.MF.up, column_title = "GO MF - FSHD up")
res.CC.dn <- simplifyGO(sim.CC.dn, column_title = "GO CC - FSHD down")
res.MF.dn <- simplifyGO(sim.MF.dn, column_title = "GO MF - FSHD down")
## computation take very long
res.BP.up <- simplifyGO(sim.BP.up, column_title = "GO BP - FSHD up")
res.BP.dn <- simplifyGO(sim.BP.dn, column_title = "GO BP - FSHD down")
```


## Software information

```{r}
sessionInfo()
```
