---
title: "Analysis of EGAD00001008337"
author: "T. Schaetzl and M. Kohl"
date: "December 2022"
output:
  rmarkdown::html_document:
    theme: united
    highlight: tango
    toc: true
---

```{r setup, include=FALSE}
## setup for the chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)
Sys.setlocale("LC_ALL", "English")
```


# Import of data

```{r}
SampleInfo <- read.csv("EGAD00001008337_SampleInfo.csv", na.strings = "N/A")
SampleInfo
cts <- read.delim(file='01_Data --- EGAS00001005782 --- RawCounts_Exprgenesonly.txt',row.names = 1)
# Filtering has already been done
cts <- as.matrix(cts)
# Removing mosaic samples
cts <- cts[,SampleInfo$Repeat_units != "Mosaic_2units"]
SampleInfo <- SampleInfo[SampleInfo$Repeat_units != "Mosaic_2units",]
#Removing asymptomatic samples
cts <- cts[,SampleInfo$SampleID != "FSHD_15_VL"]
SampleInfo <- SampleInfo[SampleInfo$SampleID != "FSHD_15_VL",]
cts <- cts[,SampleInfo$SampleID != "FSHD_23_VL"]
SampleInfo <- SampleInfo[SampleInfo$SampleID != "FSHD_23_VL",]
sort(colSums(cts)/1e6)
```


# Quality control 

## Boxplot 

```{r}
cts0 <- cts
cts0[cts0 == 0] <- min(cts0[cts0 != 0])
boxplot(cts0, log = "y", range = 0, las = 2)
```

## Similarity map

```{r}
library(MKomics)
Cor <- cor(cts, method = "spearman")
corPlot2(Cor, minCor = min(Cor), 
         #labels = paste(SampleInfo$subject),
         title = "Spearman correlation")
```


# Preprocessing

```{r}
library(limma)
library(edgeR)
dge <- DGEList(counts=cts)
group <- SampleInfo$group
logCPM <- cpm(dge, log=TRUE, prior.count=3)
```


# Quality control of preprocessed data

## Boxplot

```{r}
boxplot(logCPM, range = 0, las = 2)
```

## Similarity map

```{r}
Cor <- cor(logCPM, method = "spearman")
corPlot2(Cor, minCor = min(Cor), 
         #labels=paste(SampleInfo$subject),
         title = "Spearman correlation")
```

## Mean-SD plot

```{r}
library(vsn)
meanSdPlot(logCPM)
```

## Loading annotation data

```{r}
library(org.Hs.eg.db)
ensemblIDs <- rownames(logCPM)
entrezIDs <- mapIds(org.Hs.eg.db, keys = ensemblIDs, column = "ENTREZID", 
                    keytype = "ENSEMBL", multiVals = "first")
symbols <- mapIds(org.Hs.eg.db, keys = ensemblIDs
                  , column = "SYMBOL", 
                    keytype = "ENSEMBL", multiVals = "first")
```

## Saving of preprocessed data

```{r}
ExportData <- t(logCPM)
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "EGAD00001008337data.rds")
```


# Statistical analysis

## Moderated t-test by packages limma and MKomics

```{r}
library(limma)
library(MKomics)
res <- mod.t.test (x = logCPM, group = SampleInfo$group)
colSums(res[,c(5,6)] < 0.05)
# 7326 results for adj. p value < 5%
```

## Saving of results

```{r}
res.EGAD00001008337 <- as.data.frame(res)
res.EGAD00001008337$ENSEMBLE <- ensemblIDs
res.EGAD00001008337 <- res.EGAD00001008337[order(res.EGAD00001008337$adj.p.value),]
write.csv(res.EGAD00001008337, file = "ResultsEGAD00001008337.csv", row.names = FALSE)
save(res.EGAD00001008337, file = "Final_ResultsEGAD00001008337.RData")
```

## Heatmap 

```{r}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
## Data selection
sel <-res$adj.p.value < 0.05
selData <- logCPM[sel,]
selData <- selData - rowMeans(selData, na.rm = TRUE)
colnames(selData) <- paste(SampleInfo$SampleID)
## Color definition
col1 <- rev(brewer.pal(n = 8, name = "RdYlBu"))
col2 <- brewer.pal(n = 3, name = "Set1")[c(3,1)]
col3 <- brewer.pal(n = 3, name = "Set2")[c(3,1)]
col4 <- brewer.pal(n = 3, name = "Dark2")[c(2,4)]
mycol <- colorRamp2(seq(-1, 1, length = 128), 
                    colorRampPalette(col1)(128))
## Annotation
df <- data.frame(group = SampleInfo$group)
ha <- HeatmapAnnotation(df = df,
                        col = list(group = c("CTRL" = col2[1],
                                              "FSHD" = col2[2])))
## Drawing the heatmap
Heatmap(selData, col = mycol, name = "log2-FC", show_row_names = FALSE,
        use_raster = FALSE,
        top_annotation = ha, show_column_names = TRUE,
        column_names_gp = gpar(fontsize = 8),
        clustering_distance_columns = "euclidean",
        show_row_dend = TRUE, km = 1,
        cluster_columns = TRUE, show_column_dend = TRUE,
        column_title = "Adjusted p-value < 0.05",
        show_heatmap_legend = TRUE)
```


# Enrichment analysis

```{r}
GeneID.UP <- entrezIDs[(res$`difference in means` > 0 & 
                        res$adj.p.value < 0.05)]
GeneID.DN <- entrezIDs[(res$`difference in means` < 0 & 
                        res$adj.p.value < 0.05)]
```

## Calculation of enrichment for GO

```{r}
library(GO.db)
go <- goana(list(Up = unique(GeneID.UP), 
                 Down = unique(GeneID.DN)), 
            universe = unique(entrezIDs), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:50,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:50,]
```

## Saving of GO results

### Upregulated pathways

```{r}
go.up.EGAD00001008337 <- go.up
go.up.EGAD00001008337 <- go.up.EGAD00001008337[order(go.up.EGAD00001008337$P.Up),]
write.csv(go.up.EGAD00001008337, file = "GO.UP_EGAD00001008337.csv", row.names = FALSE)
save(go.up.EGAD00001008337, file = "GO.UP_EGAD00001008337.RData")
```

### Downregulated pathways

```{r}
go.dn.EGAD00001008337 <- go.dn
go.dn.EGAD00001008337 <- go.dn.EGAD00001008337[order(go.dn.EGAD00001008337$P.Down),]
write.csv(go.dn.EGAD00001008337, file = "GO.DN_EGAD00001008337.csv", row.names = FALSE)
save(go.dn.EGAD00001008337, file = "GO.DN_EGAD00001008337.RData")
```


# Used software

```{r}
sessionInfo()
```
