---
title: "Analysis of GSE10760"
author: "T. Schätzl and M. Kohl"
date: "December 2022"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
## setup for the chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)
Sys.setlocale("LC_ALL", "English")
```


# Import of data

The data are included in dataset GSE10760 of Gene Expression Omnibus:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE10760

The tar-files were unzipped into subfolders called GSE10760_RAW_A and GSE10760_RAW_B. 

We import the data from the CEL-files using Bioconductor package affy.

```{r}
library(affy)
rawData.A <- ReadAffy(celfile.path = "./GSE10760_RAW_A")
rawData.A
rawData.B <- ReadAffy(celfile.path = "./GSE10760_RAW_B")
rawData.B
```


# Quality control

## QC reports of package arrayQualityMetrics

```{r, eval=FALSE}
library(arrayQualityMetrics)
arrayQualityMetrics(rawData.A, outdir = "QC_A", force = TRUE, do.logtransform = TRUE)
arrayQualityMetrics(rawData.B, outdir = "QC_B", force = TRUE, do.logtransform = TRUE)
```

## Similarity maps

```{r}
library(MKomics)
COR <- cor(exprs(rawData.A), method = "spearman")
corPlot2(COR, minCor = min(COR), fontsize.axis = 8, 
         title = "Spearman correlation - U133A")
COR <- cor(exprs(rawData.B), method = "spearman")
corPlot2(COR, minCor = min(COR), fontsize.axis = 8, 
         title = "Spearman correlation - U133B")
```


# Preprocessing

We use RMA to preprocess the data.

```{r}
normData.A <- rma(rawData.A)
normData.A
normData.B <- rma(rawData.B)
normData.B
```


# Quality control of preprocessed data

## Mean-SD plots

```{r}
library(vsn)
library(ggplot2)
gg <- meanSdPlot(normData.A, plot = FALSE)
gg$gg + ggtitle("Mean-SD-Plot - U133A")
gg <- meanSdPlot(normData.B, plot = FALSE)
gg$gg + ggtitle("Mean-SD-Plot - U133B")
```

## Boxplots

```{r}
## range = 0: whiskers till most extreme data points
boxplot(exprs(normData.A), range = 0, main = "RMA - U133A", las = 2, 
        cex.axis = 0.75)
boxplot(exprs(normData.B), range = 0, main = "RMA - U133B", las = 2, 
        cex.axis = 0.75)
```

## Similarity maps

```{r}
COR.rma <- cor(exprs(normData.A), method = "spearman")
corPlot2(COR.rma, minCor = min(COR.rma), fontsize.axis = 8, 
         title = "Spearman correlation - U133A")
COR.rma <- cor(exprs(normData.B), method = "spearman")
corPlot2(COR.rma, minCor = min(COR.rma), fontsize.axis = 8, 
         title = "Spearman correlation - U133B")
```

## Import of sample information

```{r}
SampleInfo <- read.csv(file = "GSE10760_SampleInfo.csv", na.strings = "N/A")
SampleInfo
normData <- exprs(normData.A)
colnames(normData) <- 1:ncol(normData)
tmp <- exprs(normData.B)
colnames(tmp) <- 1:ncol(tmp)
normData <- rbind(normData, tmp)
```

## Saving of preprocessed data

```{r}
ExportData <- t(normData)
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE10760data.rds")
```


# Statistical analysis

## Moderated t-test by packages limma and MKomics

```{r}
library(limma)
library(MKomics)
rownames(normData) <- NULL
res <- mod.t.test (x = normData, group = SampleInfo$group)
colSums(res[,c(5,6)] < 0.05)
# 4459 results for adj. p value < 5%
```

## Results

### Loading annotation data

```{r}
library(org.Hs.eg.db)
library(hgu133a2.db)
library(hgu133b.db)
GeneID.A <- select(hgu133a2.db, keys = keys(hgu133a2.db), 
                   columns = c("ENSEMBL", "ENTREZID", "SYMBOL"))
GeneID.B <- select(hgu133b.db, keys = keys(hgu133b.db), 
                   columns = c("ENSEMBL", "ENTREZID", "SYMBOL"))
GeneID <- rbind(GeneID.A, GeneID.B)
```

### Saving of results

```{r}
res.GSE10760 <- as.data.frame(res)
res.GSE10760$PROBEID <- c(rownames(normData.A),rownames(normData.B))
res.GSE10760 <- merge(res.GSE10760, GeneID, by = "PROBEID", all.x = TRUE)
res.GSE10760 <- res.GSE10760[order(res.GSE10760$adj.p.value),]
write.csv(res.GSE10760, file = "ResultsGSE10760.csv", row.names = FALSE)
save(res.GSE10760, file = "ResultsGSE10760.RData")
```

## Heatmap 

```{r}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
## Data selection
sel <- res$adj.p.value < 0.05 
selData <- normData[sel,]
selData <- selData - rowMeans(selData, na.rm = TRUE)
colnames(selData) <- as.character(SampleInfo$subject)
## Color definition
col1 <- rev(brewer.pal(n = 8, name = "RdYlBu"))
col2 <- brewer.pal(n = 3, name = "Set1")[c(3,1)]
mycol <- colorRamp2(seq(-2, 2, length = 128), colorRampPalette(col1)(128))
## Annotation
df <- data.frame(group = SampleInfo$group)
ha <- HeatmapAnnotation(df = df,
                        col = list(group = c("control" = col2[1],
                                              "FSHD" = col2[2])))
## Drawing the heatmap
Heatmap(selData, col = mycol, name = "log2-FC", show_row_names = FALSE,
        use_raster = FALSE,
        top_annotation = ha, show_column_names = TRUE,
        column_names_gp = gpar(fontsize = 8),
        clustering_distance_columns = "pearson",
        show_row_dend = TRUE, cluster_rows = TRUE,
        cluster_columns = TRUE, show_column_dend = TRUE,
        column_title = "Adjusted p-value < 0.05",
        show_heatmap_legend = TRUE)
```


# Enrichment analysis

```{r}
ProbeIDs <- c(rownames(exprs(normData.A)), rownames(exprs(normData.B)))
keys.UP <- ProbeIDs[(res$`difference in means` > 0 & 
                            res$adj.p.value < 0.05)]
keys.DN <- ProbeIDs[(res$`difference in means` < 0 & 
                            res$adj.p.value < 0.05)]
GeneID.UP.A <- select(hgu133a2.db, keys = keys.UP, columns = "ENTREZID")
GeneID.UP.B <- select(hgu133b.db, keys = keys.UP, columns = "ENTREZID")
GeneID.UP <- rbind(GeneID.UP.A, GeneID.UP.B)
GeneID.DN.A <- select(hgu133a2.db, keys = keys.DN, columns = "ENTREZID")
GeneID.DN.B <- select(hgu133b.db, keys = keys.DN, columns = "ENTREZID")
GeneID.DN <- rbind(GeneID.DN.A, GeneID.DN.B)
```

## Calculation of enrichment for GO

```{r}
library(limma)
go <- goana(list(Up = unique(GeneID.UP$ENTREZID), 
                 Down = unique(GeneID.DN$ENTREZID)), 
            universe = unique(GeneID$ENTREZID), 
            species = "Hs")
## Print top 50
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:50,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:50,]
```

## Saving of GO results

### Upregulated pathways

```{r}
go.up.GSE10760 <- go.up
go.up.GSE10760 <- go.up.GSE10760[order(go.up.GSE10760$P.Up),]
write.csv(go.up.GSE10760, file = "GO.UP_GSE10760.csv", row.names = FALSE)
save(go.up.GSE10760, file = "GO.UP_GSE10760.RData")
```

### Downregulated pathways

```{r}
go.dn.GSE10760 <- go.dn
go.dn.GSE10760 <- go.dn.GSE10760[order(go.dn.GSE10760$P.Down),]
write.csv(go.dn.GSE10760, file = "GO.DN_GSE10760.csv", row.names = FALSE)
save(go.dn.GSE10760, file = "GO.DN_GSE10760.RData")
```


# Used software

```{r}
sessionInfo()
```

