---
title: "Analysis of GSE26061"
author: "T. Schätzl and M. Kohl"
date: "December 2022"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
## setup for the chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)
Sys.setlocale("LC_ALL", "English")
```


# Import of data

The data are included in dataset GSE26061 of Gene Expression Omnibus:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26061

The tar-file was unzipped into a subfolder called GSE26061_RAW. 
We removed all files containing data about FSHD2.

We import the data from the CEL-files using Bioconductor package oligo.

```{r}
library(oligo)
celFiles <- list.files("./GSE26061_RAW", full.names = TRUE)
rawData <- read.celfiles(celFiles)
rawData
```


# Quality control

## QC report of package arrayQualityMetrics

```{r, eval=FALSE}
library(arrayQualityMetrics)
arrayQualityMetrics(rawData, outdir = "QC", force = TRUE, do.logtransform = TRUE)
```

## Similarity map

```{r}
library(MKomics)
COR <- cor(exprs(rawData), method = "spearman")
corPlot2(COR, minCor = 0.75, fontsize.axis = 8, 
         title = "Spearman correlation")
```


# Preprocessing

We use RMA to preprocess the data.

```{r, cache = TRUE}
normData <- rma(rawData)
normData
```


# Quality control of preprocessed data

## Mean-SD plot

```{r}
library(vsn)
meanSdPlot(normData)
```

## Boxplot

```{r}
## range = 0: whiskers till most extreme data points
boxplot(exprs(normData), range = 0, main = "RMA", las = 2, 
        cex.axis = 0.75)
```

## Similarity map

```{r}
COR.rma <- cor(exprs(normData), method = "spearman")
corPlot2(COR.rma, minCor = min(COR.rma), fontsize.axis = 8, 
         title = "Spearman correlation")
```

## Import of sample information

```{r}
SampleInfo <- read.csv(file = "GSE26061_SampleInfo.csv", na.strings = "N/A")
SampleInfo
```

## Saving of preprocessed data

```{r}
ExportData <- t(exprs(normData))
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE26061data.rds")
```


# Statistical analysis

## Moderated t-test by packages limma and MKomics

```{r}
library(limma)
library(MKomics)
res <- mod.t.test (x = exprs(normData), group = SampleInfo$group)
colSums(res[,c(5,6)] < 0.05)
# no results for adj. p value < 5%
```


# Used software

```{r}
sessionInfo()
```

