---
title: "Analysis of GSE26852"
author: "T. Schätzl and M. Kohl"
date: "December 2022"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
---


```{r setup, include=FALSE}
## setup für die Chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)
Sys.setlocale("LC_ALL", "English")
```


# Import of data

The data are included in dataset GSE26852 of Gene Expression Omnibus:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26852

The tar-file was unzipped into a subfolder called GSE26852_RAW. 
We removed all files containing data about inflammatory myopathies (n=7) and dysferlinopathies (n=4).

We import the data from the probe profiles using Bioconductor packages lumi and beadarray.

```{r}
library(lumi)
Ilu.Daten <- lumiR(fileName = "GSE26852_non-normalized.txt")
Ilu.Daten <- Ilu.Daten[,1:15]
Ilu.Daten
SampleInfo <- read.csv(file = "GSE26852_SampleInfo.csv", na.strings = "N/A")
SampleInfo
## Removing of STIR- samples
Ilu.Daten <- Ilu.Daten[,SampleInfo$disease_state != "T2-STIR-"]
SampleInfo <- SampleInfo[SampleInfo$disease_state != "T2-STIR-",]
ProbeID <- read.delim(file = "GSE26852_non-normalized.txt", skip = 1)$X
library(beadarray)
Ilu.Daten2 <- readBeadSummaryData(dataFile = "GSE26852_non-normalized.txt", 
                                  skip = 1, ProbeID = "ProbeID")
Ilu.Daten2 <- Ilu.Daten2[,1:15]
Ilu.Daten2
## Removing of STIR- samples
Ilu.Daten2 <- Ilu.Daten[,SampleInfo$disease_state != "T2-STIR-"]
SampleInfo <- SampleInfo[SampleInfo$disease_state != "T2-STIR-",]
```


# Quality control

## QC report of package arrayQualityMetrics

```{r, eval=FALSE} 
library(arrayQualityMetrics)
arrayQualityMetrics(Ilu.Daten2, outdir = "QC", force = TRUE,
                    do.logtransform = TRUE)
```

## Similarity map

```{r}
library(MKomics)
COR <- cor(exprs(Ilu.Daten), method = "spearman")
corPlot2(COR, minCor = min(COR), labels = pData(Ilu.Daten)[,1], 
        title = "Spearman correlation")
```


# Preprocessing

We use RSN to preprocess the data.

```{r}
Ilu.rsn <- rsn(lumiT(Ilu.Daten, method = "log2"))
```


# Quality control of preprocessed data

## Mean-SD plot

```{r}
library(vsn)
meanSdPlot(Ilu.rsn)
```

## Boxplot 

```{r}
boxplot(Ilu.rsn, main = "rsn")
```

## Similarity map

```{r}
COR.rsn <- cor(exprs(Ilu.rsn))
corPlot2(COR.rsn, minCor = min(COR), title = "Spearman correlation")
```

## Saving of preprocessed data

```{r}
ExportData <- t(exprs(Ilu.rsn))
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE26852data.rds")
```


# Statistical analysis

## Moderated t-test by packages limma and MKomics

```{r}
library(limma)
library(MKomics)
res <- mod.t.test(x = exprs(Ilu.rsn), group = SampleInfo$group)
colSums(res[,c(5,6)] < 0.05)
# 3489 results for adj. p value < 5%
```

## Results

### Loading annotation data

```{r}
library(org.Hs.eg.db)
library(illuminaHumanv3.db)
GeneID <- select(illuminaHumanv3.db, keys = keys(illuminaHumanv3.db),
                 columns = c("ENSEMBL", "ENTREZID", "SYMBOL"))
```

### Saving of results

```{r}
res.GSE26852 <- as.data.frame(res)
res.GSE26852$PROBEID <- ProbeID
res.GSE26852 <- merge(res.GSE26852, GeneID, by = "PROBEID", all.x = TRUE)
res.GSE26852 <- res.GSE26852[order(res.GSE26852$adj.p.value),]
write.csv(res.GSE26852, file = "ResultsGSE26852.csv", row.names = FALSE)
save(res.GSE26852, file = "ResultsGSE26852.RData")
```

## Heatmap 

```{r}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
## Data selection
sel <- res$adj.p.value < 0.05 
selData <- exprs(Ilu.rsn)[sel,]
selData <- selData - rowMeans(selData, na.rm = TRUE)
colnames(selData) <- as.character(SampleInfo$group)
## Setting colors
col1 <- rev(brewer.pal(n = 8, name = "RdYlBu"))
col2 <- brewer.pal(n = 3, name = "Set1")[c(3,1)]
mycol <- colorRamp2(seq(-2, 2, length = 128), colorRampPalette(col1)(128))
## Annotation
df <- data.frame(group = SampleInfo$group)
ha <- HeatmapAnnotation(df = df,
                        col = list(group = c("control" = col2[1],
                                             "FSHD" = col2[2])))
## Drawing the Heatmap
Heatmap(selData, col = mycol, name = "log2-FC", show_row_names = FALSE,
        use_raster = FALSE,
        top_annotation = ha, show_column_names = TRUE,
        column_names_gp = gpar(fontsize = 8),
        clustering_distance_columns = "pearson",
        show_row_dend = TRUE, cluster_rows = TRUE,
        cluster_columns = TRUE, show_column_dend = TRUE,
        column_title = "Adjusted p-value < 0.05",
        show_heatmap_legend = TRUE)
```


# Enrichment analysis

```{r}
library(org.Hs.eg.db)
library(illuminaHumanv3.db)
GeneID <- select(illuminaHumanv3.db, keys = keys(illuminaHumanv3.db), 
                 columns = "ENTREZID")
keys.UP <- ProbeID[(res$`difference in means` > 0 & 
                                   res$p.value < 0.05)]
keys.DN <- ProbeID[(res$`difference in means` < 0 & 
                                   res$p.value < 0.05)]
GeneID.UP <- select(illuminaHumanv3.db, keys = keys.UP, columns = "ENTREZID")
GeneID.DN <- select(illuminaHumanv3.db, keys = keys.DN, columns = "ENTREZID")
```

## Calculation of enrichment for GO

```{r}
library(limma)
go <- goana(list(Up = unique(GeneID.UP$ENTREZID), 
                 Down = unique(GeneID.DN$ENTREZID)), 
            universe = unique(GeneID$ENTREZID), 
            species = "Hs")
go.up <- topGO(go, sort = "up", number = Inf)
go.up[1:50,]
go.dn <- topGO(go, sort = "down", number = Inf)
go.dn[1:50,]
```

## Saving of GO results

### Upregulated pathways

```{r}
go.up.GSE26852 <- go.up
go.up.GSE26852 <- go.up.GSE26852[order(go.up.GSE26852$P.Up),]
write.csv(go.up.GSE26852, file = "GO.UP_GSE26852.csv", row.names = FALSE)
save(go.up.GSE26852, file = "GO.UP_GSE26852.RData")
```

### Downregulated pathways

```{r}
go.dn.GSE26852 <- go.dn
go.dn.GSE26852 <- go.dn.GSE26852[order(go.dn.GSE26852$P.Down),]
write.csv(go.dn.GSE26852, file = "GO.DN_GSE26852.csv", row.names = FALSE)
save(go.dn.GSE26852, file = "GO.DN_GSE26852.RData")
```


# Used software

```{r}
sessionInfo()
```

