---
title: "Analysis of GSE36398"
author: "T. Schätzl and M. Kohl"
date: "December 2022"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
## setup for the chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)
Sys.setlocale("LC_ALL", "English")
```


# Import of data

The data are included in dataset GSE36398 of Gene Expression Omnibus:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE36398

The tar-file was unzipped into a subfolder called GSE36398_RAW. 
We removed all files with deltoid muscle data because there are both biceps and deltoid muscle biopsies on the same individual with an associated family control in the present data set. This is because the biceps is more affected compared with the deltoid muscle in FSHD. 

We import the data from the CEL-files using Bioconductor package oligo.

```{r}
library(oligo)
celFiles <- list.files("./GSE36398_RAW", full.names = TRUE)
rawData <- read.celfiles(celFiles)
rawData
```


# Quality control

## QC report of package arrayQualityMetrics

```{r, eval=FALSE}
library(arrayQualityMetrics)
arrayQualityMetrics(rawData, outdir = "QC", force = TRUE, do.logtransform = TRUE)
```

## Similarity map

```{r}
library(MKomics)
COR <- cor(exprs(rawData), method = "spearman")
corPlot2(COR, minCor = min(COR), fontsize.axis = 8, 
         title = "Spearman correlation")
```


# Preprocessing

We use RMA to preprocess the data.

```{r, cache = TRUE}
normData <- rma(rawData)
normData
```


# Quality control of preprocessed data

## Mean-SD plot

```{r}
library(vsn)
meanSdPlot(normData)
```

## Boxplot

```{r}
## range = 0: whiskers till most extreme data points
boxplot(exprs(normData), range = 0, main = "RMA", las = 2, 
        cex.axis = 0.75)
```

## Similarity map

```{r}
COR.rma <- cor(exprs(normData), method = "spearman")
corPlot2(COR.rma, minCor = min(COR.rma), fontsize.axis = 8, signifBar = 3,
         title = "Spearman correlation")
```

## Import of sample information

```{r}
SampleInfo <- read.csv(file = "GSE36398_SampleInfo.csv")
SampleInfo
```

Due to the different scan dates (see "GSE36398_SampleInfo.csv") a batch effect might occur. This is clearly evident from the similarity map. It is also described in the related publication (see "Supporting Information", in: Rahimov et al. 10.1073/pnas.1209508109).

## Removing of batch effects 

We use the empirical Bayes approach proposed by W.E. Johnson, C. Li, and A. Rabinovic. "Adjusting batch effects in microarray data using empirical bayes methods." Biostatistics,8(1):118–127, 2007.

```{r}
library(sva)
scanDate <- sapply(strsplit(SampleInfo$scan_date, " |T"), "[", 1)
DF <- data.frame(group = as.factor(SampleInfo$group), 
                 scanDate = scanDate)
mod.batch <- model.matrix(~ group, data=DF)
normDataBatch <- ComBat(dat = normData, 
                        batch = scanDate, mod = mod.batch, 
                        par.prior = TRUE, mean.only = TRUE)
```

### Mean-SD plot of batch corrected data

```{r}
meanSdPlot(normDataBatch)
```

### Boxplot of batch corrected data

```{r}
## range = 0: whiskers till most extreme data points
boxplot(normDataBatch, range = 0, main = "Parametric ComBat", las = 2, 
        cex.axis = 0.75)
```

### Similarity map of batch corrected data

```{r}
COR.param <- cor(normDataBatch, method = "spearman")
corPlot2(COR.param, minCor = min(COR.param), fontsize.axis = 8, 
         title = "Spearman correlation of batch corrected data", 
         signifBar = 3)
```

## Saving of preprocessed data

```{r}
ExportData <- t(normDataBatch)
ExportData <- data.frame(SampleInfo, ExportData)
saveRDS(ExportData, file = "GSE36398data.rds")
```


# Statistical analysis

## Moderated t-test by packages limma and MKomics

```{r}
library(limma)
library(MKomics)
res <- mod.t.test(x = normDataBatch, group = SampleInfo$group)
colSums(res[,c(5,6)] < 0.05)
# no results for adj. p value < 5%
```


# Used software

```{r}
sessionInfo()
```

